---
title: "MATH-M463"
subtitle: 'Homework 8'
author: "John Koo"
output: pdf_document
# output: html_document
header-includes:
- \usepackage{float}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center')
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

# Problem 1

## Part a

$\lambda(t) = t^2$  
$G(t) = \exp(-\int_0^t \lambda(u) du)$ 
$= \exp(-\int_0^t u^2 du) = \exp(-t^3 /3)$. 

$P(T > 10) = G(10) = \boxed{\exp(-1000/3)}$

## Part b

$f(t) = -\frac{dG}{dt} = -\frac{d}{dt} \exp(-t^3/3)$  
$= -(-t^2) \exp(-t^3 / 3) = \boxed{t^2 \exp(-t^3/3)}$

## Part c

$E[T^3] = \int_0^\infty t^5 \exp(-t^3 / 3) dt$

Let:  
$u = t^3$, $du = 3 t^2 dt$  
$v = -\exp(-t^3 / 3)$, $dv = t^2 \exp(-t^3/3) dt$

Then we get $-t^3 \exp(-t^3/3) |_0^\infty + 3\int_0^\infty t^2 \exp(-t^3/3)dt$

The first term goes to 0 and so we're left with 
$3\int_0^\infty t^2 \exp(-t^3/3)dt$. Then let:

$u = -t^3/3$  
$du = -t^2 dt$

Then this becomes $-3 \int_0^\infty \exp(u) du$  
$= -3 \exp(u) |_{u=0}^\infty$  
$= -3 (0 - 1) = \boxed{3}$

# Problem 2

## Part a

$P(T_2 < 16) = 1 - P(T_2 > 16) = 1 - P(N_{16} < 1)$  
$= 1 - e^{-8/3} (1 + 8/3)$ 
$= \boxed{1 - \frac{11}{3}e^{-8/3} \approx 0.745}$

## Part b

This is equivalent to an example from the book/lectures, so we have 
$m = x / \lambda = 6x$ and $x \approx 1.675$, so $m \approx \boxed{10.05}$

# Problem 3

## Part a

```{r p3_a, echo = FALSE}
library(ggplot2)
theme_set(theme_bw())

x <- seq(-3, 3, .01)
y <- dplyr::if_else(x >= 0, x ** 2 - x, -x ** 2 - x)

ggplot() + 
  geom_line(aes(x = x, y = y)) + 
  scale_x_continuous(breaks = seq(-3, 3, .5)) + 
  scale_y_continuous(breaks = seq(-6, 6))
```

## Part b

\[ \frac{dy}{dx} =
  \begin{cases}
    2x - 1 & \quad \text{if } x \geq 0\\
    -2x - 1 & \quad \text{if } x < 0
  \end{cases}
\]

## Part c

Solving for $x$ wrt $y$ for each of the cases, we get:

$$x = \frac{1 \pm \sqrt{1 \pm 4y}}{2}$$

With $(+, +)$ when $x \geq 1/2$, $(-, +)$ when $0 \leq x < 1/2$, $(+, -)$ when 
$-1/2 \leq x < 0$, and $(-, -)$ when $x < 1/2$.

# Problem 4

## Part a

When $y > 1/4$, $y(x) = x^2 - x$ and $x(y) = \frac{1 + \sqrt{1 + 4y}}{2}$. Then 
$f_Y(y) = f_X(x) / |\frac{dy}{dx}| = f_X(x) / |2x - 1|$  
$= \boxed{f_X(\frac{1 + \sqrt{1 + 4y}}{2}) \bigg/ \sqrt{1 + 4y}}$

## Part b

When $0 < y < 1/4$, $x \in [-1, 0] \cup [1, \infty)$. In terms of $y$, this 
defines 3 of the regions specified in the previous problem. Then:

$$f_Y(y) = \frac{f_X(1/2 + \sqrt{1 + 4y}/2)}{\sqrt{1 + 4y}} + 
\frac{f_X(1/2 + \sqrt{1 - 4y}/2) + f_X(1/2 - \sqrt{1 - 4y}/2)}{\sqrt{1 - 4y}}$$

# Problem 5

# Problem 6