---
title: "MATH-M463"
subtitle: 'Homework 8'
author: "John Koo"
output: pdf_document
# output: html_document
header-includes:
- \usepackage{float}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center')
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

# Problem 1

## Part a

$\lambda(t) = t^2$  
$G(t) = \exp(-\int_0^t \lambda(u) du)$ 
$= \exp(-\int_0^t u^2 du) = \exp(-t^3 /3)$. 

$P(T > 10) = G(10) = \boxed{\exp(-1000/3)}$

## Part b

$f(t) = -\frac{dG}{dt} = -\frac{d}{dt} \exp(-t^3/3)$  
$= -(-t^2) \exp(-t^3 / 3) = \boxed{t^2 \exp(-t^3/3)}$

## Part c

$E[T^3] = \int_0^\infty t^5 \exp(-t^3 / 3) dt$

Let:  
$u = t^3$, $du = 3 t^2 dt$  
$v = -\exp(-t^3 / 3)$, $dv = t^2 \exp(-t^3/3) dt$

Then we get $-t^3 \exp(-t^3/3) |_0^\infty + 3\int_0^\infty t^2 \exp(-t^3/3)dt$

The first term goes to 0 and so we're left with 
$3\int_0^\infty t^2 \exp(-t^3/3)dt$. Then let:

$u = -t^3/3$  
$du = -t^2 dt$

Then this becomes $-3 \int_0^\infty \exp(u) du$  
$= -3 \exp(u) |_{u=0}^\infty$  
$= -3 (0 - 1) = \boxed{3}$

# Problem 2

## Part a

$P(T_2 > 16) = P(N_{16} < 1)$  
$= e^{-8/3} (1 + 8/3)$ 
$= \boxed{\frac{11}{3}e^{-8/3} \approx 0.255}$

## Part b

This is equivalent to an example from the book/lectures, so we have 
$m = x / \lambda = 6x$ and $x \approx 1.675$, so $m \approx 10.05$ or 
$\boxed{10 \text{ weeks}}$.

# Problem 3

## Part a

```{r p3_a, echo = FALSE}
library(ggplot2)
theme_set(theme_bw())

x <- seq(-3, 3, .01)
y <- dplyr::if_else(x >= 0, x ** 2 - x, -x ** 2 - x)

ggplot() + 
  geom_line(aes(x = x, y = y)) + 
  scale_x_continuous(breaks = seq(-3, 3, .5)) + 
  scale_y_continuous(breaks = seq(-6, 6))
```

## Part b

\[ \frac{dy}{dx} =
  \begin{cases}
    2x - 1 & \quad \text{if } x \geq 0\\
    -2x - 1 & \quad \text{if } x < 0
  \end{cases}
\]

## Part c

Solving for $x$ wrt $y$ for each of the cases, we get:

$$x = \frac{1 \pm \sqrt{1 \pm 4y}}{2}$$

With $(+, +)$ when $x \geq 1/2$, $(-, +)$ when $0 \leq x < 1/2$, $(+, -)$ when 
$-1/2 \leq x < 0$, and $(-, -)$ when $x < 1/2$.

# Problem 4

## Part a

When $y > 1/4$, $y(x) = x^2 - x$ and $x(y) = \frac{1 + \sqrt{1 + 4y}}{2}$. Then 
$f_Y(y) = f_X(x) / |\frac{dy}{dx}| = f_X(x) / |2x - 1|$  
$= \boxed{f_X(\frac{1 + \sqrt{1 + 4y}}{2}) \bigg/ \sqrt{1 + 4y}}$

## Part b

When $0 < y < 1/4$, $x \in [-1, 0] \cup [1, \infty)$. In terms of $y$, this 
defines 3 of the regions specified in the previous problem. Then:

$$\boxed{
  f_Y(y) = \frac{f_X(1/2 + \sqrt{1 + 4y}/2)}{\sqrt{1 + 4y}} + 
  \frac{f_X(1/2 + \sqrt{1 - 4y}/2) + f_X(1/2 - \sqrt{1 - 4y}/2)}{\sqrt{1 - 4y}}
}$$

# Problem 5

## Part a

```{r, echo = FALSE}
ggplot() + 
  geom_segment(aes(x = 1, xend = 0, y = 0, yend = 0), 
               arrow = arrow()) + 
  geom_segment(aes(x = 1, xend = 2, y = 1/6, yend = 1/6)) + 
  geom_segment(aes(x = 2, xend = 3, y = 1/3, yend = 1/3)) + 
  geom_segment(aes(x = 3, xend = 4, y = 1/2, yend = 1/2)) + 
  geom_segment(aes(x = 4, xend = 5, y = 2/3, yend = 2/3)) + 
  geom_segment(aes(x = 5, xend = 6, y = 5/6, yend = 5/6)) + 
  geom_segment(aes(x = 6, xend = 7, y = 1, yend = 1), 
               arrow = arrow()) + 
  geom_point(aes(x = seq(6), y = seq(1/6, 1, 1/6))) + 
  labs(x = 'x', y = 'F(x)') + 
  scale_x_continuous(breaks = seq(0, 7))
```

## Part b

$F_{max}(x) = P(X_1 \leq x)^4$, so 

\[ F_{max}(x) =
  \begin{cases}
    0 & \quad \text{if } x < 1\\
    \lfloor \frac{x}{6} \rfloor^4 & \quad \text{if } 1 \leq x < 6\\
    1 & \quad \text{otherwise}
  \end{cases}
\]

## Part c

$F_{min}(x) = (1 - P(X_1 \leq x))^4$, so 

\[ F_{min}(x) = 
  \begin{cases}
    0 & \quad \text{if } x < 1\\
    1 - (1 - \lfloor \frac{x}{6} \rfloor)^4 & \quad \text{if } 1 \leq x < 6\\
    1 & \quad \text{otherwise}
  \end{cases}
\]

# Problem 6

## Part a

$f(x) = \frac{d}{dx} \frac{4}{\pi} \arctan(x)$  
$= \boxed{\frac{4}{\pi} \frac{1}{x^2 + 1}}$

## Part b

$\boxed{
  f_{(3)}(x) = 5 \frac{4}{\pi} \frac{1}{x^2+1} \binom{4}{2} 
  (\frac{4}{\pi} \arctan(x))^2 (1 - \frac{4}{\pi} \arctan(x))^2
}$