---
title: "STAT-S632"
subtitle: 'Assignment 2'
author: "John Koo"
# output: pdf_document
output: html_document
urlcolor: blue
header-includes:
- \usepackage{float}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center', 
                      fig.height = 5, 
                      fig.width = 5)
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

```{r packages_etc}
# packages, etc.
import::from(magrittr, `%>%`, `%<>%`)
dp <- loadNamespace('dplyr')
library(ggplot2)
import::from(GGally, ggpairs)
theme_set(theme_bw())
```

# ELM 2.1

```{r problem_setup}
# get the data
wbca.df <- faraway::wbca
summary(wbca.df)
```

## Part a

```{r a_0}
ggplot(wbca.df) + 
  geom_point(aes(x = BNucl, y = Class), alpha = .1)
```

### i

A scatterplot does not work well for `Class ~ BNucl` because the relationship 
is not regressive. `Class` is a quantitativve variable with only two levels, 
so a scatterplot will have a hard time showing any sort of trend between 
it and a predictor.

### ii

```{r a_ii}
wbca.df %<>% dp$mutate(Class = as.character(Class))

ggplot(wbca.df) + 
  geom_boxplot(aes(x = Class, group = Class, fill = Class, y = BNucl)) + 
  scale_fill_brewer(palette = 'Set1') + 
  labs(fill = NULL)

table(dp$filter(wbca.df, Class == '0')$BNucl)
table(dp$filter(wbca.df, Class == '1')$BNucl)
```

A boxplot reveals that most of the values of `BNucl` for `Class == 0` are 
$geq 5$, whereas most of the `BNucl` values for `Class = 1` are 1. 

### iii

```{r a_iii}
wbca.df %>% 
  dp$mutate(Class = as.numeric(Class)) %>% 
  ggplot() + 
  geom_jitter(aes(x = BNucl, y = Class), 
              height = .01, width = .2, alpha = .1)
```

We can see that there is a concentration at `BNucl = 1` when `Class = 1` and 
at `BNucl = 10` when `Class = 0`.