---
title: "STAT-S632"
subtitle: 'Assignment 3'
author: "John Koo"
output: pdf_document
# output: html_document
urlcolor: blue
header-includes:
- \usepackage{float}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center', 
                      fig.height = 5, 
                      fig.width = 5)
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

```{r packages_etc}
# packages, etc.
import::from(magrittr, `%>%`, `%<>%`)
dp <- loadNamespace('dplyr')
library(ggplot2)
import::from(GGally, ggpairs)
theme_set(theme_bw())
```

# Exercise 3.2

```{r problem_setup}
# get the data
turtles.df <- faraway::turtle %>% 
  dp$mutate(turtles = male + female, 
            prop.male = male / turtles)
summary(turtles.df)
```

## Part a

```{r a, fig.height = 3}
# p vs temp
ggplot(turtles.df) + 
  geom_point(aes(x = temp, y = prop.male)) + 
  labs(x = 'temperature [C]', y = 'proportion male')

# eta vs temp
ggplot(turtles.df) + 
  geom_point(aes(x = temp, y = log(prop.male / (1 - prop.male)))) + 
  labs(x = 'temperature [C]', y = expression(eta))
```

We can see that the proportion of male turtles increases with temperature. 
However, it does not look like a logit is a good fit on these data. In 
particular, when we transform the $p_i$s to $\eta_i$s 
($\eta = \log \frac{p}{1-p}$), we do not get a linear relationship between 
$\eta$ and temperature.

## Part b

```{r b}
# fit the model
binom.mod <- glm(cbind(male, female) ~ temp, 
                 data = turtles.df, family = binomial)
summary(binom.mod)

pchisq(binom.mod$deviance, binom.mod$df.residual, lower = FALSE)
pchisq(summary(binom.mod)$null.deviance, 
       summary(binom.mod)$df.null, 
       lower = FALSE)
```

The residual deviance suggests that the model is not a good fit for these data.

## Part c

```{r c}
# number of observations for each sub-sample
summary(turtles.df$turtles)
```

Using our rule of thumb of $m_i > 5$, we can say that the data are not sparse.

## Part d

```{r d}
# half normal plot
faraway::halfnorm(residuals(binom.mod))
```

Based on the half-normal plot, we have no reason to believe that the data 
contain outliers.

## Part e

```{r e}
# empirical log plot
ggplot(turtles.df) + 
  geom_point(aes(x = temp, y = log((male + .5) / (turtles - male + .5)))) + 
  labs(x = 'temperature [C]', y = 'empirical logits') + 
  stat_smooth(aes(x = temp, y = log((male + .5) / (turtles - male + .5))), 
              method = 'lm')
```

Plotting the empirical logits vs the temperature suggests that the relationship 
is not linear. It appears that a concave curve is more appropriate. A 
transformation or higher order term might be appropriate.

## Part f

```{r f}
# model with quadratic term
quad.binom.mod <- glm(cbind(male, female) ~ temp + I(temp ** 2), 
                      data = turtles.df, family = binomial)
# wald tests
summary(quad.binom.mod)

# check for model fit
pchisq(quad.binom.mod$deviance, quad.binom.mod$df.residual, lower = FALSE)

# LR test
anova(binom.mod, quad.binom.mod, test = 'Chi')
```

Using a significance level of $\alpha = 0.05$, we can say that the quadratic 
term is significant and the model (kinda) fits the data. 

## Part g

```{r g}
# sample variance
var(turtles.df$male)

# sample variance assuming binomial
turtles.df %>% 
  dp$group_by(temp) %>% 
  dp$summarise(var.male = sum(male) * sum(female) / sum(turtles)) %>% 
  dp$ungroup() %>% 
  .$var.male %>% 
  sum()  # assuming independence

# dispersion parameter estimate
s2 <- sum(residuals(quad.binom.mod, type = 'pearson') ** 2) / 
  quad.binom.mod$df.residual

# F tests
drop1(quad.binom.mod, scale = s2, test = 'F')
```

Comparing the sample variance to the sample variance assuming a binomial model, 
as well as the estimate for the dispersion parameter, suggests that there is 
some overdispersion. 

## Part h

```{r h}
# aggregated data
agg.turtles.df <- turtles.df %>% 
  dp$group_by(temp) %>% 
  dp$summarise_all(sum) %>% 
  dp$ungroup() 

# model
combin.binom.mod <- glm(cbind(male, female) ~ temp, 
                        data = agg.turtles.df, family = binomial)
# wald tests
summary(combin.binom.mod)

# check for model fit
pchisq(combin.binom.mod$deviance, combin.binom.mod$df.residual, lower = FALSE)
```

As before, the residual deviance suggests that the model does not fit the data. 
The summary outputs for both this model and the model from part (b) are 
identical. 