---
title: "STAT-S632"
subtitle: 'Assignment 6'
author: "John Koo"
output: pdf_document
# output: html_document
fontsize: 9pt 
# geometry: "left=1.5cm,right=1.5cm,top=1.5cm,bottom=1.5cm"
urlcolor: blue
header-includes:
- \usepackage{float}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center', 
                      fig.height = 4, 
                      fig.width = 6)
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

```{r setup2}
import::from(magrittr, `%>%`, `%<>%`)
library(ggplot2)
import::from(lme4, lmer, VarCorr, ranef)
import::from(RLRsim, exactRLRT)

theme_set(theme_bw())
```

# Problem 1

```{r p1_setup}
denim.df <- faraway::denim
summary(denim.df)
```

## Part a

```{r p1_a}
ggplot(denim.df) + 
  geom_violin(aes(x = supplier, y = waste, fill = supplier)) + 
  geom_point(aes(x = supplier, y = waste), alpha = .3) + 
  scale_fill_brewer(palette = 'Set1') + 
  guides(fill = FALSE)
```

## Part b

```{r p1_b}
fixed.mod <- lm(waste ~ supplier, data = denim.df)
summary(fixed.mod)
drop1(fixed.mod, test = 'F')
```

`supplier` is not significant.

## Part c

```{r p1_c}
rand.eff.mod <- lmer(waste ~ 1 + (1 | supplier), data = denim.df)
summary(rand.eff.mod)
```

$\beta = \mu \in \mathbb{R}^1$

$X \in \mathbb{R}^{95}$ (`nrow(denim.df)`)

$y, \epsilon \in \mathbb{R}^{95}$ as well

$\gamma \in \mathbb{R}^5$, or the number of levels of `supplier`

$Z = \begin{bmatrix}
  z_1 \\ & z_2 \\ & & z_3 \\ & & & z_4 \\ & & & & z_5 
\end{bmatrix}$ where $\dim(z_i) = |\{x : x = \alpha_i\}|$, or the number of 
elements in level $i$. So $Z \in \mathbb{R}^{95 \times 5}$.

## Part d

### Parametric bootstrapping

```{r p1_d_bootstrap, cache = TRUE}
confint(rand.eff.mod, method = 'boot')
```

Since the lower bound of the interval is 0, we cannot say that the variance 
of `supplier` is significant. We can also see this from the $p$-value.

```{r p1_d_boostrap_2}
exactRLRT(rand.eff.mod)
```

### LR test

```{r p1_d_lr}
mle.mod <- lmer(waste ~ 1 + (1 | supplier), data = denim.df, REML = FALSE)
pchisq(2 * (logLik(mle.mod) - logLik(fixed.mod)), 1, lower.tail = FALSE)
```

We get a $p$-value of 1.

## Part e

```{r p1_e}
ranef(rand.eff.mod)

lattice::dotplot(ranef(rand.eff.mod, condVar = TRUE))
```

# Problem 2

```{r p2_setup}
breaking.df <- faraway::breaking
summary(breaking.df)
```

## Part a

```{r p2_a}
ggplot(breaking.df, 
       aes(y = y, x = operator, colour = day, shape = supplier)) + 
  geom_point() + 
  scale_colour_brewer(palette = 'Set1')
```

```{r p2_a_2, fig.width = 16}
ggplot(breaking.df, 
       aes(y = y, x = operator, colour = day)) + 
  geom_point() + 
  scale_colour_brewer(palette = 'Set1') + 
  facet_wrap(~ supplier, labeller = 'label_both', nrow = 1)

ggplot(breaking.df, 
       aes(y = y, x = operator, colour = supplier)) + 
  geom_point() + 
  scale_colour_brewer(palette = 'Set1') + 
  facet_wrap(~ day, labeller = 'label_both', nrow = 1)
```

There appears to be a suggestion of differences among suppliers but not among 
operators or days.

## Part b

$X \in \mathbb{R}^{16 \times 4}$ (16 rows $\times$ 4 factor levels)

$\beta \in \mathbb{R}^4$ (4 factor levels)

$y, \epsilon \in \mathbb{R}^{16}$ (16 rows)

$\gamma \in \mathbb{R}^7$ (1 intercept + $(3-1)$ from factor 1 + $(3-1)$ from 
factor 2)

$Z \in \mathbb{R}^{16 \times 7}$, or the number of rows $\times$ $\dim(\gamma)$

## Part c

```{r p2_c}
fixed.mod <- lm(y ~ day + operator + supplier, data = breaking.df)
summary(fixed.mod)
drop1(fixed.mod, test = 'F')

anova(fixed.mod, lm(y ~ supplier, data = breaking.df))
fixed.mod <- lm(y ~ supplier, data = breaking.df)
summary(fixed.mod)
```

It appears that `supplier` is the only significant covariate.

## Part d

```{r p2_d}
mixed.mod <- lmer(y ~ supplier + (1 | operator) + (1 | day), data = breaking.df)
summary(mixed.mod)
```

The experiment was conducted to select a supplier, so it is the effect of 
interest. 

Supplier C has the highest breaking point, as we saw in the fixed effect model 
and the plots.

## Part e

```{r p2_e}
mixed.op.mod <- lmer(y ~ supplier + (1 | operator), data = breaking.df)
mixed.day.mod <- lmer(y ~ supplier + (1 | day), data = breaking.df)

exactRLRT(mixed.op.mod, mixed.mod, mixed.day.mod)
exactRLRT(mixed.day.mod, mixed.mod, mixed.op.mod)
```

Neither are significant.