---
title: "STAT-S632"
subtitle: 'Assignment 7'
author: "John Koo"
output: pdf_document
# output: html_document
fontsize: 9pt 
# geometry: "left=1.5cm,right=1.5cm,top=1.5cm,bottom=1.5cm"
urlcolor: blue
header-includes:
- \usepackage{float}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center', 
                      fig.height = 4, 
                      fig.width = 6)
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

# Problem 1

```{r p1_setup}
library(ggplot2)
import::from(magrittr, `%>%`)
library(lme4)
import::from(pbkrtest, KRmodcomp)

theme_set(theme_bw())

rat.test.df <- faraway::ratdrink
summary(rat.test.df)
```

## Part a

```{r p1_a}
ggplot(rat.test.df) + 
  geom_line(aes(x = weeks, y = wt, group = subject, colour = treat)) + 
  scale_colour_brewer(palette = 'Set1')

ggplot(rat.test.df) + 
  geom_point(aes(x = weeks, y = wt, group = subject, colour = treat)) + 
  scale_colour_brewer(palette = 'Set1') + 
  scale_fill_brewer(palette = 'Set1') + 
  stat_smooth(aes(x = weeks, y = wt, colour = treat, fill = treat))
```

```{r p1_a_2, fig.width = 12}
ggplot(rat.test.df) + 
  geom_line(aes(x = weeks, y = wt, group = subject)) + 
  facet_wrap(~treat)
```

Weight increases over time, but it appears that the rats treated with thlouracil 
gained less weight.

## Part b

```{r p1_b}
mixed.mod <- lmer(wt ~ weeks * treat + (weeks | subject), data = rat.test.df)
summary(mixed.mod)
```

* Intercept - The average weight at week 0 for the baseline (control) group
* `weeks:thiouracil` - On average, each week the rats in the thiouracil group 
gained 9.4 units less in weight compared to the baseline (control) group.
* $\hat{\sigma}_{intercept}$ - Since this is greater than 
$\hat{\sigma}_\epsilon$, the subject effect is not negligible. We can also 
compute the ICC, which is $\approx 0.632$. 

## Part c

The $D$ matrix has the form $diag(d, d, ..., d)$ where each of the 27 (one for 
each subject) $d$s is a $2 \times 2$ variance-covariance matrix of the form 
$d = \begin{bmatrix} 
  \sigma_1^2 & \sigma_{12} \\ \sigma_{12} & \sigma_2^2
\end{bmatrix}$. So $D \in \mathbb{R}^{54 \times 54}$.

## Part d

```{r p1_d}
week.mod <- lmer(wt ~ weeks + (weeks | subject), data = rat.test.df)
KRmodcomp(mixed.mod, week.mod)
```

The treatment effect is significant, but we can also see whether the interaction 
term is.

```{r p1_d_2}
no.inter.mod <- lmer(wt ~ weeks + treat + (weeks | subject), 
                     data = rat.test.df)
KRmodcomp(mixed.mod, no.inter.mod)
```

And we can see that the interaction term is also significant.

## Part e

```{r p1_e, fig.width = 4}
diag.df <- ggplot2::fortify(mixed.mod)
summary(diag.df)

ggplot(diag.df) +
  stat_qq(aes(sample = .resid))

qqnorm(ranef(mixed.mod)$subject$`(Intercept)`)
qqnorm(ranef(mixed.mod)$subject$weeks)
```

```{r p1_e_2}
ggplot(diag.df) + 
  geom_point(aes(x = .fitted, y = .resid), alpha = .2) + 
  geom_abline(slope = 0, colour = 'red')
```

We don't have too much evidence of non-normality or heteroscedacity of the 
residuals. There also isn't much evidence that the random effects are not 
normally distributed. We also don't see any outliers.

## Part f

```{r p1_f, cache = TRUE}
confint(mixed.mod, method = 'boot')
```

Based on the confidence intervals, the random effects are significant. However, 
looking at the fixed effects, we can see that since both the marginal and 
interaction terms for `thyroxine` have confidence intervals that contain 0, 
they may not differ significantly from the control group. 