---
title: "STAT-S676"
subtitle: 'Assignment 1'
author: "John Koo"
# output: pdf_document
output: html_document
urlcolor: blue
header-includes:
- \usepackage{float}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center', 
                      fig.height = 5, 
                      fig.width = 5)
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

# Problem 1

AIC is given by $-2 \log(L(\hat{\theta}|y)) + 2K$ where $L$ is the likelihood 
function. We are given that $Y \sim \mathcal{N}(X \beta, \sigma^2 I)$. Then 
$\varepsilon = Y - X \beta \sim \mathcal(0, \sigma^2 I)$. Therefore, the pdf of 
$\varepsilon$ is a multivariate normal 
$f(\varepsilon) = \big( \frac{1}{2 \pi \sigma^2} \big)^{n/2} 
e^{-\frac{1}{2 \sigma^2} \varepsilon^T \varepsilon}$. 
$\varepsilon^T \varepsilon = RSS = n \hat{\sigma}^2 \approx n \sigma^2$. 
(We sub in the estimate for the parameter.) Then we get 
$L = \big( \frac{1}{2 \pi \sigma^2} \big)^{n/2} e^{-\frac{n}{2}}$. Plugging 
this in for $L$, we get:

$$AIC = -2 \log 
\bigg( \Big( \frac{1}{2 \pi \sigma^2} \Big)^{n/2} e^{-\frac{n}{2}} \bigg) + 
2K$$
$$= -2 
\bigg(-\frac{n}{2} \log(2 \pi) - \frac{n}{2} \log(\sigma^2) - \frac{n}{2} \bigg)
+ 2K$$
$$= n \log(2 \pi) + n \log(\sigma^2) + n + 2K$$

We can ignore the two terms $n \log(2\pi)$ and $n$ since they are fixed for some 
particular dataset.

# Problem 2

# Problem 3

TIC is defined by $-2 \log L + 2 tr(J(\theta) I(\theta^{-1}))$. 
$L(\theta|y) = g(y|\theta)$

# Problem 4

```{r packages_etc}
# packages, etc.
import::from(magrittr, `%>%`, `%<>%`)
dp <- loadNamespace('dplyr')

load('~/dev/stats-hw/stat-s676/diabetes2.Rdata')
```