---
title: "STAT-S631"
subtitle: 'Exam 1'
author: "John Koo"
# output: pdf_document
output: html_document
header-includes:
- \usepackage{float}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center')
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

# Statement

On my honor, I have not had any form of communicationabout this exam with any 
other individual (including other students, teaching assistants, instructors, 
etc.).

Signed: *John Koo*

```{r packages_and_data}
# packages
import::from(magrittr, `%>%`)
dp <- loadNamespace('dplyr')
library(ggplot2)
import::from(GGally, ggpairs)

# plotting stuff
theme_set(theme_bw())

# read data
city.df <- read.table('~/dev/stats-hw/stat-s631/takehome1.txt')
```

# Question 1

```{r q1}
N <- nrow(city.df)
prop.crime.above.3200 <- nrow(dp$filter(city.df, crime > 3200)) / N

ggplot(city.df) + 
  geom_histogram(aes(x = crime), 
                 fill = 'white', 
                 colour = 'black', 
                 binwidth = 200) + 
  labs(x = 'crime rate [per 100,000]')

qqnorm(city.df$crime)
```

From the data, the estimated probability that a randomly selected city will have 
a crime rate higher than 3,200 per 100,000 people is `r prop.crime.above.3200`.

The histogram of `crime` seems to suggest that the data are normally 
distributed. The histogram is a symmetric, unimodal bell curve. A QQ-plot 
confirms this.

# Question 2

```{r q2}
ggpairs(city.df, lower = list(continuous = 'smooth'))

city.df %>% 
  dp$transmute(log.population = log(population), 
               log.nonwhite = log(nonwhite), 
               log.density = log(density), 
               crime) %>% 
  ggpairs(lower = list(continuous = 'smooth'))
```

From the pairwise scatterplots, if `crime` is the response variable, a log 
transformation is appropriate for all three of the covariates.

```{r q2_2}
model.2 <- lm(crime ~ log(population), data = city.df)
summary(model.2)
```

## Part a

From the scatterplot of `crime` vs `population`, there are no obvious violations 
of our assumptions. The trend looks linear and there is an even number of 
points to either side of the OLS line. 

## Part b

