---
title: 'STATS-S631'
subtitle: 'Assignment 1'
author: 'John Koo'
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      fig.pos = 'H', 
                      fig.align = 'center')
```

# Problem 1

$S = \{\text{set of all UN members}\}$

$X : S \rightarrow \mathbb{R}$, maps a female life expectancy from each country

$X$ is a random variable since $\forall r \in \mathbb{R}$ there is a probability 
that $X(s) = r$ for some $s \in S$.

# Problem 2

a. $\{s \in S : X(s) \leq 80\}$

b. $\{s \in S : X(s) = 75\}$

c. $\{s \in S : X(s) \in [65, 70]\}$

# Problem 3

a. $P(S \leq 80)$

b. $P(S \leq 75) - P(S \leq 74)$

c. $P(S \leq 70) - P(S \leq 64)$

# Problem 4

First, load the data as a variable

```{r load_data}
un.df <- alr4::UN11
```

Then compute the probabilities.

```{r compute_probs}
part.a <- nrow(un.df[un.df$lifeExpF <= 80, ]) / nrow(un.df)

part.b <- nrow(un.df[un.df$lifeExpF == 75, ]) / nrow(un.df)

part.c <- 
  nrow(un.df[un.df$lifeExpF <= 70 & un.df$lifeExpF >= 65, ]) / nrow(un.df)
```

(Results rounded to 3 decimal places)

a. `r round(part.a, 3)`

b. `r round(part.b, 3)`

c. `r round(part.c, 3)`

# Problem 5

```{r plots, fig.height = 3, fig.width = 4}
import::from(ggplot2, 
             ggplot, theme_set, aes, geom_density, labs, stat_ecdf)
theme_set(ggthemes::theme_base())

ggplot(un.df) + 
  geom_density(aes(x = lifeExpF)) + 
  labs(x = 'Life Expectancy [years]')

ggplot(un.df) + 
  stat_ecdf(aes(x = lifeExpF)) + 
  labs(x = 'Life Expectancy [years]',
       y = 'Cumulative density')
```

The distribution does not appear normal. It's not symmetric and there is a clear
skew to the left. It might be possible to approximate it as the sum of two 
normal distributions. We can verify this by computing the skewness of the
distribution:

```{r skew}
print(moments::skewness(un.df$lifeExpF))
```

Since the skewness is much closer to -1 than to 0, we can say that a symmetric
distribution, such as the normal distribution, is not a good approximation for
these data.

Although it's pretty clear at this point that the population cannot be 
approximated well with a normal distribution, if it weren't clear, we could do a
few quick tests.

One such test is to check the quantile-quantile plot. Since the points do not
fall in a straight line, we can say that a normal distribution cannot
approximate the data:

```{r qqplot, fig.height = 3, fig.width = 3}
qqnorm(un.df$lifeExpF)
```