---
title: "STAT-S631"
subtitle: 'Assignment 11'
author: "John Koo"
output: pdf_document
# output: html_document
urlcolor: blue
header-includes:
- \usepackage{float}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center', 
                      fig.height = 5, 
                      fig.width = 5)
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

```{r packages_etc}
dp <- loadNamespace('dplyr')
import::from(magrittr, 
             `%>%`, `%<>%`)
library(ggplot2)
theme_set(theme_bw())
import::from(GGally, 
             ggpairs)
import::from(car, 
             Anova, boxCox, bcPower, 
             powerTransform, invResPlot, 
             invTranEstimate, invTranPlot)
import::from(effects, 
             effect, Effect)
import::from(miscTools, 
             rSquared)
```

# Problem 1

[From ALR 8.2]

```{r p1_setup}
stopping.df <- alr4::stopping
```

## Part 1

```{r p1_1}
lin.mod <- lm(Distance ~ Speed, data = stopping.df)

invResPlot(lin.mod, lambda = seq(-5, 5))
boxCox(lin.mod)
```

The optimal $\lambda$ is 0.485. However, $\lambda = .5$ is in the 95\% confidence 
interval, and it is the only integer value in the interval. So we will use 
$\lambda = .5$ for this problem.

## Part 2

```{r p1_2}
invTranPlot(Distance ~ Speed, data = stopping.df, 
            lambda = seq(-1, 1), optimal = FALSE)
```

From the scatterplot with fitted lines, we can see that none of these values 
of $\lambda$ fit the data very well. $\lambda = 0$ or $-1$ do not lie near the 
points, and $\lambda = 1$ fails to capture the curvature, resulting in a 
pattern in the residuals.

## Part 3

```{r p1_3}
invTranPlot(Distance ~ Speed, data = stopping.df, lambda = 2)
invTranEstimate(stopping.df$Speed, stopping.df$Distance)
```

The optimal $\lambda$ and $\lambda = 2$ produce very similar results. In 
addition, 2 is in the 95\% CI for the optimal $\lambda$ (minimizing the 
log-likelihood).

## Part 4

```{r p1_4}
hald.mod <- lm(Distance ~ Speed + I(Speed ** 2), 
               weights = I(Speed ** -2), 
               data = stopping.df)
summary(hald.mod)

power.mod <- lm(sqrt(Distance) ~ Speed, 
                data = stopping.df)
summary(power.mod)

stopping.df %<>%
  dp$mutate(hald.pred = predict(hald.mod, newdata = stopping.df), 
            power.pred = predict(power.mod, newdata = stopping.df) ** 2, 
            hald.resid = Distance - hald.pred, 
            power.resid = Distance - power.pred)
ggplot(stopping.df) +
  geom_point(aes(x = Speed, y = Distance))  + 
  geom_line(aes(x = Speed, y = hald.pred, colour = 'Hald')) + 
  geom_line(aes(x = Speed, y = power.pred, colour = 'power transform')) + 
  theme(legend.position = 'bottom') + 
  labs(colour = NULL) + 
  scale_colour_brewer(palette = 'Set1')

mean(stopping.df$hald.resid ** 2)
mean(stopping.df$power.resid ** 2)
rSquared(stopping.df$Distance, stopping.df$hald.resid)
rSquared(stopping.df$Distance, stopping.df$power.resid)
```

# Problem 2

[From ALR 8.6]

```{r p2_setup}
wool.df <- car::Wool
```

## Part 1

```{r p2_1, cache = TRUE, fig.height = 3}
ggpairs(wool.df)

summary(wool.df)

wool.df %>% 
  dp$select(len, amp, load) %>% 
  table()

dim(wool.df)
```

The values for `len`, `amp`, and `load` consist of just 3 values each. Each 
triple is unique, which matches the number of rows of the data frame ($3^3$). The values are evenly spaced out. 

## Part 2

```{r p2_2, fig.width = 5, fig.height = 5}
wool.df %<>% 
  dp$mutate(len = as.factor(len), 
            amp = as.factor(amp), 
            load = as.factor(load))

factor.2.mod <- lm(cycles ~ len * amp + len * load + amp * load, 
                   data = wool.df)
summary(factor.2.mod)
Anova(factor.2.mod)
plot(effect('len:amp', factor.2.mod), multiline = TRUE, ci.style = 'bars')
```

For a significance level of $\alpha = 0.05$, we reject the null hypothesis that 
the coefficients for the `len` and `amp` interaction terms is 0. 

## Part 3

```{r p2_3_1}
factor.1.mod <- lm(cycles ~ len + amp + load, data = wool.df)
summary(factor.1.mod)

anova(factor.2.mod, factor.1.mod)
```

The ANOVA test confirms the text's assertion.

```{r p2_3_2}
invResPlot(factor.1.mod)
boxCox(factor.1.mod)
summary(powerTransform(factor.1.mod))
```

The best value of $\lambda$ (the one that maximizes the log-likelihood) is 
-0.1005. However, 0 is within the 95\% confidence interval, so we cannot say 
that -0.1005 is better than 0 (for $\alpha = 0.05$). So we will select 
$\lambda = 0$.

## Part 4

```{r p2_4_1}
factor.1.log.mod <- lm(log(cycles) ~ len + amp + load, data = wool.df)
factor.2.log.mod <- lm(log(cycles) ~ len * amp + len * load + amp * load, 
                       data = wool.df)
anova(factor.2.log.mod, factor.1.log.mod)
```

From the ANOVA test, we fail to reject the null hypothesis that all of the
coefficients for the interaction terms is 0.

```{r p2_4_2}
plot(Effect(c('len', 'amp'), factor.2.log.mod, 
            transformation = list(link = log, inverse = exp)), 
     multiline = TRUE, 
     ci.style = 'bars')
plot(Effect(c('len', 'amp'), factor.2.log.mod, 
            transformation = list(link = log, inverse = exp)), 
     multiline = TRUE, 
     axes = list(y = list(type = 'response', lab = 'cycles')), 
     ci.style = 'bars')
```

The confidence interval increases with `cycles`. 