---
title: "STAT-S631"
subtitle: 'Assignment 7'
author: "John Koo"
output: pdf_document
# output: html_document
header-includes:
- \usepackage{float}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      fig.pos = 'H', 
                      fig.align = 'center')
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

```{r packages}
dp <- loadNamespace('dplyr')
import::from(magrittr, `%>%`, `%<>%`)
library(ggplot2)
theme_set(theme_bw())
import::from(GGally, ggpairs)
```

# Problem 1

[From ALR 4.1]

```{r q1_setup, cache = TRUE}
# set up data
bgsgirls.df <- alr4::BGSgirls %>% 
  dp$mutate(ave = (WT2 + WT9 + WT18) / 3, 
            lin = WT18 - WT2, 
            quad = WT2 - 2 * WT9 + WT18)

# pairwise scatterplot
bgsgirls.df %>% 
  dp$transmute(ave, lin, quad, BMI18) %>% 
  ggpairs(lower = list(continuous = 'smooth'))

# fit model
model.1 <- lm(BMI18 ~ ave + lin + quad, data = bgsgirls.df)
summary(model.1)

# also model from 4.1
model.4.1 <- lm(BMI18 ~ WT2 + WT9 + WT18, data = bgsgirls.df)
summary(model.4.1)
```

From the scatterplot matrix, we know that while both `ave` and `lin` correlate 
fairly strongly and positively with `BMI18`, they also correlate with each 
other. On the other hand, `quad` doesn't seem to correlate very strongly with 
any of the variables. 

From the model summary, we can see that only the intercept term and the 
coefficient estimate for `lin` are significant at reasonable significance levels 
(e.g., $\alpha = 0.01$ or $0.05$). This is consistent with what we observed in 
the scatterplots: `quad` has no correlation with `BMI18`, and although both 
`ave` and `lin` correlate with `BMI18`, they correlate with each other, so we 
expect some unexpected behavior. In this case, the model fit `BMI18` on `lin` 
and not on `ave`. 

Comparing the results here to the example in Section 4.1, we can see that 
$\hat{\beta}_0$ is the same here as it is in the example. This is expected, 
since the regressors are all from the same set of three predictors, and although 
they end up being at different scales, this can just be adjusted by the 
estimates for the $\beta$s. Summary statistics such as the residual standard 
error, $R^2$, and $F$-statistic are identical, which is, again, expected, since 
the regressors are derived from the same set of predictors. 

# Problem 2

[From ALR 4.2]

```{r p2}
transact.df <- car::Transact %>% 
  dp$mutate(a = (t1 + t2) / 2, 
            d = t1 - t2)

m1 <- lm(time ~ t1 + t2, data = transact.df)
summary(m1)

m2 <- lm(time ~ a + d, data = transact.df)
summary(m2)

m3 <- lm(time ~ t2 + d, data = transact.df)
summary(m3)

m4 <- lm(time ~ t1 + t2 + a + d, data = transact.df)
summary(m4)
```

## Part 1

Coefficients are omitted when the model matrix isn't full rank. Since `a` and 
`d` are linear combinations of `t1` and `t2`, the model matrix of `m4`, which 
has five columns, isn't of rank 5. 

## Part 2

All of the models have the same $\hat{\beta}_0$ (including standard error), 
$R^2$, residuals (and therefore, RMSE), $F$-statistic, and degrees of freedom 
(since two of the four regressors are omitted in `m4`). However, they do not 
have the same $\hat{\beta}_1$ or $\hat{\beta}_2$, even when the corresponding 
regressors are the same. 

## Part 3

`d` is a linear combination of `t1` and `t2`. Therefore, `m2` and `m3` are 
equivalent models. 

Starting with model 2:

$$E[Y | a, d] = \beta_0 + \beta_{21} a + \beta_{22} d$$
$$= \beta_0 + \beta_{21}\frac{t_1 + t_2}{2} + \beta_{22} (t_1 - t_2)$$
$$= \beta_0 + (\frac{\beta_{21}}{2} + 
\beta_{22}) t_1 + (\frac{\beta_{21}}{2} - \beta_{22}) t_2$$

Therefore, for model 3, $\beta_{31} = \frac{\beta_{21}}{2} + \beta_{22}$ and 
$\beta_{32} = \frac{\beta_{21}}{2} - \beta_{22}$.

# Problem 3

[From ALR 4.6 and 4.7]

$$\log{(\hat{\text{fertility}})} = 1.501 - 0.01 \text{pctUrban}$$
So ...

$$\hat{\text{fertility}} = \exp{(1.501 - .01 \text{pctUrban})}$$
$$= 4.486 e^{-0.01 \text{pctUrban}}$$

Then for one unit increase in `pctUrban`, `fertility` is multiplied by 
$4.487 e^{-0.01}$ (on average). 

```{r p3}
un11.df <- alr4::UN11
model.4.7 <- lm(log(fertility) ~ log(ppgdp) + lifeExpF, data = un11.df)
summary(model.4.7)
```

The model here is:

$$E[\log{Y} | x_1, x_2] = \beta_0 + \beta_1 \log{x_1} + \beta_2 x_2$$

This is equivalent to (after some algebra):

$$E[Y | x_1, x_2] = \gamma_0 x_1^{\beta_1} e^{\beta_2 x_2}$$

Where $\gamma_0 = e^{\beta_0}$.

Then if `ppgdp`, which is $x_1$ in this case (and replacing our parameters with 
estimates) we can verify that a 25\% increase in $x_1$ would yield:

$$\frac{\hat{\gamma}_0 (1.25 x_1)^{\hat{\beta}_1} e^{\hat{\beta}_2 x_2})}
{\hat{\gamma}_0 x_1^{\hat{\beta}_1} e^{\hat{\beta}_2 x_2}}$$

$$= \frac{(1.25 x_1) ^ {\hat{\beta}_1}}{x_1 ^ {\hat{\beta}_1}}$$

$$= 1.25 ^ {\hat{\beta}_1} \approx 0.9855$$

And $1 - 0.9855 = 0.014497$