---
title: "STAT-S631"
subtitle: 'Assignment 6'
author: "John Koo"
# output: pdf_document
output: html_document
header-includes:
- \usepackage{float}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      fig.pos = 'H', 
                      fig.align = 'center')
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

```{r packages}
dp <- loadNamespace('dplyr')
import::from(magrittr, `%>%`, `%<>%`)
library(ggplot2)
theme_set(theme_bw())
import::from(GGally, ggpairs)
```

# Problem 1

[From ALR 4.1]

```{r q1_setup, cache = TRUE}
bgsgirls.df <- alr4::BGSgirls %>% 
  dp$mutate(ave = (WT2 + WT9 + WT18) / 3, 
            lin = WT18 - WT2, 
            quad = WT2 - 2 * WT9 + WT18)

bgsgirls.df %>% 
  dp$transmute(ave, lin, quad, BMI18) %>% 
  ggpairs(lower = list(continuous = 'smooth'))

model.1 <- lm(BMI18 ~ ave + lin + quad, data = bgsgirls.df)

summary(model.1)
```

From the scatterplot matrix, we know that while both `ave` and `lin` correlate 
fairly strongly and positively with `BMI18`, they also correlate with each 
other. On the other hand, `quad` doesn't seem to correlate very strongly with 
any of the variables. 

From the model summary, we can see that only the intercept term and the 
coefficient estimate for `lin` are significant at reasonable significance levels 
(e.g., $\alpha = 0.01$ or $0.05$).

# Problem 2

[From ALR 4.2]

# Problem 3

[From ALR 4.6 and 4.7]