---
title: "STAT-S631"
subtitle: 'Assignment 8'
author: "John Koo"
output: pdf_document
# output: html_document
urlcolor: blue
header-includes:
- \usepackage{float}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center', 
                      fig.width = 5, 
                      fig.height = 5)
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

```{r load_data, results = 'asis'}
dp <- loadNamespace('dplyr')
import::from(magrittr, `%>%`, `%<>%`)
import::from(readr, read_table2)
import::from(xtable, xtable)
library(ggplot2)
theme_set(theme_bw())

robey.df <- read.table('~/dev/stats-hw/stat-s631/Robey.txt') %>% 
  dp$mutate(country = rownames(.))

head(robey.df) %>% 
  xtable() %>% 
  print(include.rownames = FALSE)
```

`tfr` (total fertility rate) makes the most sense as a response variable since 
`contraceptors` (percent of married women using contraception) would directly 
affect this. 

# Part a

```{r a_viz}
robey.df %>%
  ggplot() +
  geom_boxplot(aes(x = region, group = region, y = tfr, fill = region)) +
  scale_fill_brewer(palette = 'Set1') + 
  theme(legend.position = 'bottom')
```

From a plot of fertility rate vs. region, we might expect that there is a 
somewhat significant difference between Africa and the Near East while there 
isn't much of a significant difference between Asia and Latin America. 

```{r a_mod}
region.mod <- lm(tfr ~ region, data = robey.df)
summary(region.mod)
```

From the model summary, we can see that the significance value between 
Africa (the baseline in the model) vs the Near East is 0.0544. If we set our 
$\alpha = 0.05$ or $0.01$, then we would fail to reject the null hypothesis, 
although the value is still very close to $0.05$ and below $0.1$, another common 
significance level. 

Comparing Asia and Latin America is a bit more difficult. One thing we can do 
is perform the following hypothesis test:

$H_0 : \beta_1 = \beta_2$  
$H_A : \beta_1 \neq \beta_2$

Then our test statistic is a $t$ value with degrees of freedom  $n - p - 1$:
$$t_{n - p - 1} = 
\frac{\hat{\beta}_2 - \hat{\beta}_3}{\sqrt{a^T \hat{\sigma}^2 (X^T X)^{-1} a}}$$

where $SE_i$ is the sample standard error for the $i$^th^ coefficient (written 
this way for brevity).

```{r beta_2_3_test}
coefs <- summary(region.mod)$coefficients
a <- c(0, 1, -1, 0)
t.stat <- 
  (coefs['regionAsia', 'Estimate'] - coefs['regionLatin.Amer', 'Estimate']) / 
  sqrt(t(a) %*% vcov(region.mod) %*% a)

# p-value
pt(t.stat, region.mod$df.residual) * 2
```

So if $\alpha = 0.05$ or $0.01$, we would fail to reject the null hypothesis, 
but again, if our significance level was just 10\%, we would reject the null 
hypothesis.

# Part b

The intercept term $\hat{\beta}_0$ is the average fertility rate of the baseline 
region, Africa. So on average, the fertility rate of African countries is 
around $5.856$. 

The rest of the coefficients are relative to the baseline. So Asia's average 
fertility rate is $2.316$ less than Africa's, Latin America's average fertility 
rate is $1.806$ less than Africa's, and the Near East's average fertility rate 
is $1.056$ less than Africa's.

$E[Y | Africa] = \hat{\beta}_0 = 5.856$  
$E[Y | Asia] = \hat{\beta}_0 + \hat{\beta}_1 = 5.856 - 2.316 = 3.54$  
$E[Y | Latin America] = \hat{\beta}_0 + \hat{\beta}_2 = 5.856 - 1.806 = 4.05$  
$E[Y | Near East] = \hat{\beta}_0 + \hat{\beta}_3 = 5.856 - 1.056 = 4.8$

# Part c

```{r c}
plot(effects::Effect('region', region.mod, confidence.level = .95))
```

The above plot shows the sample mean and 95\% confidence interval of fertility 
rate for each region. Going back to part a, the sample mean fertility rate of 
Africa is inside the 95\% confidence interval of the fertility rate of the Near 
East (and vice versa), and the same is true of Asia and Latin America. 

# Part d

```{r d, fig.width = 8}
ggplot(robey.df) + 
  geom_point(aes(x = contraceptors, y = tfr, colour = region)) + 
  stat_smooth(method = 'lm', 
              aes(x = contraceptors, y = tfr, 
                  fill = region, colour = region), 
              se = FALSE) + 
  scale_colour_brewer(palette = 'Set1') + 
  scale_fill_brewer(palette = 'Set1') + 
  labs(x = 'contraception rate [%]', 
       y = 'fertility rate [children per woman]') + 
  theme(legend.position = 'bottom')

ggplot(robey.df) + 
  geom_point(aes(x = contraceptors, y = tfr, colour = region)) + 
  stat_smooth(method = 'lm', 
              aes(x = contraceptors, y = tfr, 
                  fill = region, colour = region), 
              alpha = .1) + 
  scale_colour_brewer(palette = 'Set1') + 
  scale_fill_brewer(palette = 'Set1') + 
  labs(x = 'contraception rate [%]', 
       y = 'fertility rate [children per woman]') + 
  theme(legend.position = 'bottom')

ggplot(robey.df) + 
  geom_point(aes(x = contraceptors, y = tfr, colour = region)) + 
  stat_smooth(method = 'lm', 
              aes(x = contraceptors, y = tfr, 
                  fill = region, colour = region)) + 
  scale_colour_brewer(palette = 'Set1') + 
  scale_fill_brewer(palette = 'Set1') + 
  labs(x = 'contraception rate [%]', 
       y = 'fertility rate [children per woman]') + 
  theme(legend.position = 'bottom') + 
  facet_wrap(~ region, nrow = 1)
```

We can nitpick a bit, but overall, it appears that there's no reason to believe 
that there is a significant difference in slopes or intercepts among the 
regions. Especially once we add the standard errors, they all overlap. 

# Part e

```{r e}
full.mod <- lm(tfr ~ contraceptors * region, data = robey.df)
summary(full.mod)
```

As expected, the `region` terms (differences in the intercept among regions) 
and the interaction terms (differences in the slope among regions) are all 
not significant, although we can't say this just from the individual $t$-tests.

Again, the baseline region is Africa. On average, if no country in Africa used 
contraception, a fertility rate of 6.833\% is expected (although we can't 
exactly say this since contraception rate of 0 isn't in the data for any 
country). Similarly, the average fertility rates of Asia, Latin America, and the 
Near East are `r round(full.mod$coefficients[1] + full.mod$coefficients[3], 3)`%, 
`r round(full.mod$coefficients[1] + full.mod$coefficients[4], 3)`%, and 
`r round(full.mod$coefficients[1] + full.mod$coefficients[5], 3)`% respectively, 
given that the contraceptive use rate is 0. 

Moving onto the slopes, the average fertility rate in Africa decreases by 
0.054 per 1% increase in contraception use. For Asia, this value is 
`r round(full.mod$coefficients[2] + full.mod$coefficients[6], 3)`, for 
Latin America, it's 
`r round(full.mod$coefficients[2] + full.mod$coefficients[7], 3)` and for the 
Near East, it's 
`r round(full.mod$coefficients[2] + full.mod$coefficients[8], 3)`.

# Part f

First, we can build a few more models:

```{r f_models}
contraceptor.mod <- lm(tfr ~ contraceptors, data = robey.df)
summary(contraceptor.mod)

no.interact.mod <- lm(tfr ~ contraceptors + region, data = robey.df)
summary(no.interact.mod)

slopes.mod <- lm(tfr ~ contraceptors:region, data = robey.df)
summary(slopes.mod)

anova(contraceptor.mod, no.interact.mod, full.mod)
anova(region.mod, no.interact.mod, full.mod)
anova(slopes.mod, full.mod)
anova(contraceptor.mod, slopes.mod)
```

Just from the individual model summaries, we might be tempted to say 
`slopes.mod`, the model for which each region has its own slope but they all 
share the same intercept, is the best model, since all of the coefficients are 
significant. However, we can't take all of the individual $t$-tests together as 
one summary, and from the $F$-tests, it appears that this model is not 
significantly different from the model that disregards region altogether and 
just regresses on contraception use rate. 

# Part g

```{r g}
ggplot(robey.df) + 
  stat_smooth(method = 'lm', colour = 'black', 
              aes(x = contraceptors, y = tfr)) + 
  geom_point(aes(x = contraceptors, y = tfr, colour = region)) + 
  scale_colour_brewer(palette = 'Set1') + 
  labs(x = 'contraception rate [%]', 
       y = 'fertility rate [children per woman]') + 
  theme(legend.position = 'bottom')

summary(contraceptor.mod)
```

# Part h