---
title: "STAT-S631"
subtitle: 'Assignment 9'
author: "John Koo"
output: pdf_document
# output: html_document
urlcolor: blue
header-includes:
- \usepackage{float}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center')
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

# Problem 1

[From ALR 5.14]

```{r p1_setup}
dp <- loadNamespace('dplyr')
import::from(magrittr, `%>%`, `%<>%`)
import::from(car, Anova)
import::from(ggplot2, 
             ggplot, 
             geom_point, stat_smooth, 
             aes, 
             scale_colour_brewer, scale_fill_brewer, scale_x_log10, 
             labs,
             theme_set, theme_bw)
theme_set(theme_bw())

bgsall.df <- alr4::BGSall %>% 
  dp$mutate(sex = dp$if_else(Sex == 0, 'male', 'female'))
```

## Part 1

```{r p1_1}
ggplot(bgsall.df) + 
  geom_point(aes(x = HT9, y = HT18, shape = sex, colour = sex)) + 
  stat_smooth(aes(x = HT9, y = HT18), method = 'lm', colour = 'black') + 
  stat_smooth(aes(x = HT9, y = HT18, colour = sex, fill = sex), 
              method = 'lm') + 
  scale_colour_brewer(palette = 'Set1') + 
  scale_fill_brewer(palette = 'Set1')
```

From the scatterplot, there's a fair amount of separation between the sexes, and 
it appears that a model containing both `HT9` and `sex` but not an interaction 
between the two would be the most appropriate.

## Part 2

```{r p1_2}
ht9.model <- lm(HT18 ~ HT9, data = bgsall.df)
parallel.model <- lm(HT18 ~ HT9 + sex, data = bgsall.df)
full.model <- lm(HT18 ~ HT9 * sex, data = bgsall.df)

Anova(full.model)
```

From the $F$-test (which is the same as a $t$-test in this case since the factor 
only has two levels), we obtain a $p$-value of 0.0875, which is significant at 
the $\alpha = .1$ level but not at $\alpha = .05$. On the other hand, the 
$p$-value for the intercept term is significant. This test compares the model 
with just `HT9` vs. the model with both `HT9` and `sex` without the interaction 
term. 

## Part 3

```{r p1_3}
summary(parallel.model)

confint(parallel.model, 'sexmale')
```

# Problem 2

We are given:

$X = [X_1 | X_2]$  
$H = X (X^T X)^{-1} X^T$  
$H_R = X_1 (X_1^T X_1)^{-1} X_1^T$

## Part a

#### Show $H_R X_1 = X_1$

$H_R X_1 = X_1 (X_1^T X_1)^{-1} X_1^T X_1$ 
$= X_1 (X_1^T X_1)^{-1} (X_1^T X_1) = X_1$

#### Show $H X_1 = X_1$

Consider $H X$. We know that $HX = X$, so we can say:

$$HX = H [x_0, x_1, x_2, ..., x_p]$$
$$= [H x_0, H x_1, ..., H x_p]$$

Where $x_i$ is the $i$^th^ column vector of $X$.

But then $HX = X = [x_0, ..., x_p]$. Therefore:

$$[H x_0, ..., H x_p] = [x_0, ..., x_p]$$
$$\implies H x_i = x_i$$

Then if we consider $H X_1$:

$$ H X_1 = H [x_0, ..., x_{p + 1 - q}]$$
$$= [H x_0, ..., H x_{p + 1 - q}]$$
$$= [x_0, ..., x_{p + 1 - q}]$$
$$= H_1$$

#### Show $H H_R = H_R$

$H H_R = H \big(X_1 (X_1^T X_1)^{-1} X_1^T \big)$ 
$= (HX_1) (X_1^T X_1)^{-1} X_1^T = X_1 (X_1^T X_1)^{-1} X_1^T = H_R$

## Part b

#### Show $H - H_R$ is symmetric

$H - H_R$ is symmetric iff $H - H_R = (H - H_R)^T$.

We also know that $H$ and $H_R$ are symmetric. 

Therefore, $(H - H_R)^T = H^T - H_R^T = H - H_R$.

#### Show $H - H_R$ is idempotent

$H - H_R$ is idempotent iff $(H - H_R)^2 = H - H_R$

We know that $H$ and $H_R$ are idempotent. 

Therefore:

$$(H - H_R) (H - H_R) = HH - H H_R - H_R H H_R H_R$$
$$= H - H_R - H_R H + H_R$$
$$= H - H_R H$$

Consider that $H$ and $H_R$ are symmetric and $H H_R = H_R$. Therefore, 
$H_R = H_R^T = (H H_R)^T = H_R^T H^T = H_R H$  
$\implies H_R = H_R H$.

Therefore:

$$H - H_R H = H - H_R$$
$$\implies (H - H_R)^2 = H - H_R$$

## Part c

$$\frac{SSreg}{\sigma^2} = \frac{RSS_R - RSS_F}{\sigma^2}$$
$$= \frac{Y^T (I - H_R) Y - Y^T (I - H) Y}{\sigma^2}$$
$$= \frac{Y^T (H - H_R) Y}{\sigma^2}$$
$$= \frac{(Y - X_1 \hat{\beta}_1)^T (H - H_R) (Y - X_1 \hat{\beta}_1)}{\sigma^2}$$

We know that $Y - X_1 \hat{\beta}_1 \sim \mathcal{N}(0, \sigma^2 (I - H_R))$.
Furthermore, we know that 
$rank(H - H_R) = rank(H) - rank(H_R) = p + 1 - (p + 1 - q) = q$ (assuming $H$ 
and $H_R$ are full rank).

Then $\frac{SSreg}{\sigma^2} \sim \chi^2_q$ if 
$(\frac{H - H_R}{\sigma^2})(\sigma^2 (I - H_R)) = (H - H_R)(I - H_R)$ is 
idempotent. But 
$(H - H_R)(I - H_R) = H - H H_R - H_R + H_R H_R = H - H_R - H_R + H_R = H - H_R$ 
which we already know to be idempotent. Therefore, 

$$\frac{SSreg}{\sigma^2} \sim \chi^2_q$$

## Part d

$$\hat{\sigma}^2 = \frac{RSS}{n - p - 1}$$
$$= Y^T \frac{I - H}{n - p - 1} Y$$

So we have to show:

$$\bigg(\frac{H - H_R}{\sigma^2} \bigg)
\bigg(\sigma^2 (I - H) \bigg)
\bigg(\frac{I - H}{n - p - 1} \bigg)
= 0$$

We know that the product of the first two components is $H - H_R$. Therefore, 

$$\bigg(\frac{H - H_R}{\sigma^2} \bigg)
\bigg(\sigma^2 (I - H) \bigg)
\bigg(\frac{I - H}{n - p - 1} \bigg)
= (H - H_R)(I - H) \frac{1}{n - p - 1}$$
$$= (H - H - H_R + H_R) \frac{1}{n - p - 1}$$
$$= 0$$

## Part e

We know that $\frac{SSreg}{\sigma^2} \sim \chi^2_q$ and 
$\frac{RSS}{\sigma^2} \sim \chi^2_{n - p - 1}$. We also know that they are 
independent. Then we know that 
$\frac{\frac{SSreg}{\sigma^2} / q}{\frac{RSS}{\sigma^2} / (n - p - 1)}$ 
$= \frac{SSreg / q}{RSS / (n - p - 1)} \sim F_{q, n - p - 1}$

# Problem 3

[From ALR 6.4]

```{r p3_setup}
un11.df <- alr4::UN11 %>% 
  dp$mutate(country = rownames(.))

h0.model <- lm(lifeExpF ~ log(ppgdp) + group:log(ppgdp), data = un11.df)
summary(h0.model)

ha.model <- lm(lifeExpF ~ group * log(ppgdp), data = un11.df)
summary(ha.model)
```

## Part 1

The full model is one in which each `group` has its own intercept and slope. 
If the null hypothesis is true, each `group` has its own slope but the 
intercepts are not unique. 

## Part 2

```{r p3_2}
anova(h0.model, ha.model)

ggplot(un11.df) + 
  geom_point(aes(x = ppgdp, y = lifeExpF, colour = group)) + 
  labs(x = 'per capita GDP', y = 'female life expectancy') + 
  stat_smooth(aes(x = ppgdp, y = lifeExpF, colour = group, fill = group), 
              method = 'lm') + 
  scale_fill_brewer(palette = 'Set1') + 
  scale_colour_brewer(palette = 'Set1') + 
  scale_x_log10()
```

We fail to reject the null hypothesis at the $\alpha = .05$ level but do reject 
the null hypothesis at the $\alpha = .1$ level. If we go by $\alpha = .05$, then 
we would say that we do not have enough evidence to say that each `group` has 
its own intercept if we also say that each `group` has its own slope. In 
particular, we can say that if the null hypothesis is true, the probability of 
obtaining our data or something that's more extreme than our data compared to 
the null hypothesis is around 0.0556. Looking at the scatterplot, it appears 
that it would make more sense to try a model with both terms but without the 
interaction term.

## Additional part

$H_0$ : $\beta_{02} - \beta_{03} = 14$ and $\beta_{12} + \beta_{13} = .2$

(ere I'm assuming the first index refers to the intercept vs. slope while the 
second refers to `group`. In that case, we are looking to see if the `group` 
"other" has an intercept that is 14 units higher than that of the `group` 
"africa" and that the sum of the slopes for `group`s "other" and "africa" is 0.2 
units. 

```{r p3_additional}
# construct L
L <- rbind(c(0, 1, -1, 0, 0, 0), 
           c(0, 0, 0, 0, 1, 1))

# construct c
c.vec <- c(14, .2)

# find beta.hat
beta.vec <- ha.model$coefficients

# find the variance of beta.hat
V <- vcov(ha.model)

# rows of L
q. <- nrow(L)

# compute the F-statistic
F.stat <- 
  t(L %*% beta.vec - c.vec) %*% 
  solve(L %*% V %*% t(L)) %*% 
  (L %*% beta.vec - c.vec) / 
  q.

# find p-value
1 - pf(F.stat, q., ha.model$df.residual)
```

So for sensible values of $\alpha$, we would fail to reject the null hypothesis 
as specified above. 