---
title: "STAT-S620"
subtitle: 'Assignment 6'
author: "John Koo"
output: pdf_document
# output: html_document
urlcolor: blue
header-includes:
- \usepackage{float}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center', 
                      fig.height = 5, 
                      fig.width = 5)
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

```{r setup2}
import::from(magrittr, `%>%`)
```

# 4.1.3

```{r p4_1_3}
dplyr::data_frame(age = c(18, 19, 20, 21, 25), 
                  count = c(20, 22, 4, 3, 1)) %>% 
                  {sum(.$age * .$count) / sum(.$count)}
```

Then $E[X] = \boxed{18.92}$

# 4.1.4

```{r p4_1_4}
'the girl put on her beautiful red hat' %>% 
  strsplit(' ') %>% 
  unlist() %>% 
  sapply(nchar) %>% 
  mean()
```

The average number of words is $\boxed{3.75}$.

# 4.1.8

$E[XY] = \int_{x=0}^1 dx \int_{y=0}^x xy \times 12 y^2 dy$  
$= \int_0^1 x dx \int_0^x 12y^3 dy$  
$= \int_0^1 x (3y^4 |_0^x) dx$  
$= \int_0^1 3x^5 dx$  
$= 3/6 = \boxed{1/2}$

# 4.2.3

Since $X_i \sim Unif(0, 1)$, $E[X_i] = 1/2$ and $E[X_i^2] = \int_0^1 x^2 dx =$ $
1/3$.

$E[(X_1 - 2 X_2 + X_3)^2] =$ 
$E[X_1^2 - 4 X_1 X_2 + 2 X_1 X_3 + 4 X_2^2 - 4 X_2 X_3 + X_3^2]$ 
$= E[X_1^2] + 4 E[X_2^2] + E[X_3^2] - 4 E[X_1 X_2] + 2 E[X_1 X_3] - 4 E[X_2 X_3]$
$= 1/3 + 4/3 + 1/3 - 4 E[X_1 X_2] + 2 E[X_1 X_3] - 4 E[X_2 X_3]$
$= \boxed{2 - 4 E[X_1 X_2] + 2 E[X_1 X_3] - 4 E[X_2 X_3]}$

If we assume that each $X_i$ are independent, then $E[X_i X_j] = E[X_i] E[X_j]$ 
for $i \neq j$. Then the above becomes $2 - 4 (1/2)^2 + 2 (1/2)^2 - 4(1/2)^2$ 
$= \boxed{1/2}$.

# 4.2.4

We want to find $E[XY]$. Since $X$ and $Y$ are independent, this is equivalent 
to $E[X] E[Y]$. From the previous problem, we can say that $E[X] = 1/2$. And 
$E[Y] = (9 + 5) / 2 = 7$. Therefore, $E[X] E[Y] = \boxed{7/2}$.

# 4.3.2

From 4.1.4, we saw that $E[X] = 3.75$, so $E[X]^2 = 14.0625$. Then we need to 
find $E[X^2]$:

```{r p4_3_2}
'the girl put on her beautiful red hat' %>% 
  strsplit(' ') %>% 
  unlist() %>% 
  sapply(nchar) %>% 
  sapply(function(x) x ** 2) %>% 
  mean()
```

So $Var(X) = 18.25 - 14.0625 = \boxed{4.1875}$.

# 4.4.3

From a formula from class, we have $E[(X - \mu)^3] =$ 
$\mu_3 - 3 \mu_1 \mu_2 + 2 \mu_1^3 = \boxed{1}$

# 4.9.7

$E[Y - X] = \int_0^6 dy \int_0^y (y - x) \frac{x}{36} dx$ 
$= \frac{1}{36} \int_0^6 dy \int_0^y xy - x^2 dx$ 
$= \frac{1}{36} \int_0^6 (\frac{x^2 y}{2} - \frac{x^3}{3} |_0^y) dy$ 
$= \int_0^6 \frac{y^3}{216} dy$ 
$= \frac{y^4}{864} |_0^6 = \boxed{\frac{3}{2}}$

# 5.2.6

Using $E[X] = np$ if $X \sim Binom(n, p)$:

$E[A + B + C] = E[A] + E[B] + E[C] = 3/8 + 5/4 + 2/2 = \boxed{21/8}$