---
title: 'S721 HW4'
author: 'John Koo'
output: pdf_document
# output: html_document
# geometry: "left=1cm,right=1cm,top=1cm,bottom=1.5cm"
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center', 
                      fig.height = 3, 
                      fig.width = 6, 
                      fig.dpi = 300)
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

# Problem 1.39

## Part a

If $A$ and $B$ are mutually exclusive, then $A \cap B = \emptyset$ 
$\implies P(A \cap B) = 0$. 

Assume that $A$ and $B$ are independent. Then $P(A \cap B) = P(A) P(B)$. But the 
left hand side is equal to 0, which means $P(A) = 0$ or $P(B) = 0$, which is a 
contradiction.

## Part b

If $A$ and $B$ are independent, then $P(A \cap B) = P(A) P(B)$.

Assume that $A$ and $B$ are mutually exclusive. Then $P(A \cap B) = 0$. But this 
is equal to $P(A) P(B)$. Therefore, $P(A) = 0$ or $P(B) = 0$, which is a 
contradiction.

# Problem 1.40

## Part b

$P(A^c \cap B) = P(A^c \mid B) P(B) = (1 - P(A \mid B)) P(B)$. Since $A$ and $B$ 
are independent, $P(A \mid B) = P(A)$. Therefore, 
$(1 - P(A \mid B) P(B) = (1 - P(A)) P(B) = P(A^c) P(B)$.

## Part c

$P(A^c \cap B^c) = P(B^c \mid A^c) P(A^c) = (1 - P(B \mid A^c)) P(A^c)$
$= (1 - P(B)) P(A^c)$ (from part (b))  
$= P(B^c) P(A^c)$

# Problem 2.1.b

$Y = 4X + 3 \implies X = \frac{Y - 3}{4}$

Since $X > 0$, $Y > 3$.

$f_Y(y) = f_X(x(y)) \big|\frac{dx}{dy}\big|$  
$= 7 \exp \big(-\frac{7(y - 3)}{4} \big) \frac{1}{4}$  
$= \frac{7}{4} \exp \big(-\frac{7}{4} (y - 3) \big)$

Then if we integrate,  
$\int_{y = 3}^\infty f_Y(y) dy$  
$= \int_3^\infty \frac{7}{4} \exp \big(-\frac{7}{4} (y - 3) \big) dy$  
$= -\exp \big( -\frac{7}{4} (y - 3) \big) \bigg|_3^\infty$  
$= -\exp(-\infty) + \exp(0) = 1$

# Problem 2.2.b

$Y = -\log X \implies X = \exp(-Y) \implies \frac{dx}{dy} = - \exp(-Y)$

Since $0 < x < 1$, $0 < y < \infty$.

$f_Y(y) = f_X(x(y)) \big| \frac{dx}{dy} \big|$  
$= \frac{(n + m + 1)!}{n! m!} \exp(-ny) (1 - \exp(-y))^m \big| -\exp(-y) \big|$  
$= \frac{(n + m + 1)!}{n! m!} \exp \big( -y(n + 1) \big) 
\big(1 - \exp(-y) \big)^m$

# Problem 2.5

Using Figure 2.1.1, we can divide the domain into four regions, $(0, \pi / 2]$, 
$(\pi / 2, \pi]$, $(\pi, 3 \pi / 2]$, and $(3 \pi / 2, 2 \pi)$. In the first and 
third regions, $x = \arcsin(\sqrt{y})$, in the second region, 
$x = \pi - \arcsin(-\sqrt{y})$, and in the fourth region, 
$x = 2 \pi - \arcsin(-\sqrt{y})$. In any case, 
$\big| \frac{dx}{dy} \big| = \frac{1}{2 \sqrt{y(1 - y)}}$.

Then $f_Y(y) = f_X(x(y)) \big| \frac{dx}{dy} \big| \times 4$ (four regions)  
$= \frac{1}{2 \pi} \frac{1}{2 \sqrt{y (1 - y)}} \times 4$  
$= \frac{1}{\pi \sqrt{y (1 - y)}}$

# Not from text

$F(x \mid \theta) = \int_0^x \frac{1}{B(\theta, 1)} t^{\theta - 1} dt$  
$= \frac{x^\theta}{\theta B(\theta, 1)}$

Then $F^{-1}(y) = \big(\theta B(\theta, 1) y\big)^{1 / \theta}$, which we will 
use to draw our samples.

Note that $B(\theta, 1) = \frac{\Gamma(\theta) \Gamma(1)}{\Gamma(\theta + 1)}$
$= \frac{\Gamma(\theta)}{\Gamma(\theta) \theta} = \frac{1}{\theta}$. Then our 
pdf is simply $f(x \mid \theta) = \theta x^{\theta - 1}$ and our cdf is 
simply $F(x \mid \theta) = x^\theta$ (so $F^{-1}(y) = y^{1 / \theta}$).

```{r fig.height = 6, fig.width = 3}
library(ggplot2)
import::from(magrittr, `%>%`)

theme_set(theme_bw())

# as specified by the problem
theta <- 2.5
n.vector <- c(10, 100, 1000)

draw.beta <- function(n, theta) {
  # draw from the beta distribution with parameters theta and 1
  Y <- runif(n)
  Y ^ (1 / theta)
}

# ground truth data
dx <- 1e-3
x <- seq(0, 1, by = dx)
y <- theta * x ^ (theta - 1)

# construct plots for each n
plots <- lapply(n.vector, function(n) {
  X <- draw.beta(n, theta)
  ggplot() + 
    geom_histogram(aes(x = X, y = ..density..), 
                   colour = 'black', fill = 'white') + 
    geom_density(aes(x = X)) + 
    geom_line(aes(x = x, y = y), linetype = 2) +
    labs(title = paste('n =', n)) %>% 
    return()
})

do.call(gridExtra::grid.arrange, plots)
```

In the above plots, the solid line is the estimated density from the sample and 
the dashed line is the ground truth. We can see that as $n$ grows, we get a 
better approximation to the ground truth.