---
title: "STAT-S632"
subtitle: 'Experimental Design'
# output: pdf_document
output: html_document
urlcolor: blue
header-includes:
- \usepackage{float}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center', 
                      fig.height = 5, 
                      fig.width = 8)
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

# Two factors with no replication

The model is:

$$y_{ij} = \mu + \alpha_i + \beta_j + (\alpha \beta)_{ij} + \epsilon_{ij}$$

where $i = 1, ..., I$ and $j = 1, ..., J$

$\alpha$ is the effect of the first factor at level $i$ and $\beta$ is the 
effect of the second factor at level $j$.

R example:

```{r}
# library(faraway)
library(ggplot2)
library(car)

theme_set(theme_bw())

data(composite, package = 'faraway')

ggplot(composite) + 
  geom_line(aes(x = laser, y = strength, group = tape, colour = tape))

ggplot(composite) + 
  geom_line(aes(x = tape, y = strength, group = laser, colour = laser))

lmod <- lm(strength ~ laser + tape + laser:tape, data = composite)
summary(lmod)

lmod <- lm(strength ~ laser + tape, data = composite)
summary(lmod)
```

## Randomized block design

Given ...

* one treatment factor $\tau$ with $t$ levels
* one blocking factor $\beta$ with $r$ levels
* one observation on each treatment in each block
* limited ability to determine interactions between treatment and group 
(e.g., not enough observations)

The model is:

$$y_{ij} = \mu + \tau_i + \beta_j + \epsilon_{ij}$$

where $i = 1, ..., t$ and $j = 1, ..., r$

$\mu$ is the overall mean, $\tau_i$ is the treatment effect, and $\beta$ is 
the blocking effect

Can check for treatment effects and blocking effects (but not interactions)

Blocking effects are a feature of experimental units---cannot regain degrees of 
freedom if blocking is not significant

```{r}
data(oatvar, package = 'faraway')
summary(oatvar)
xtabs(yield ~ variety + block, oatvar)

ggplot(oatvar) + 
  geom_line(aes(x = block, y = yield, group = variety, colour = variety)) + 
  scale_colour_brewer(palette = 'Set1')

ggplot(oatvar) + 
  geom_line(aes(x = variety, y = yield, group = block, colour = block)) + 
  scale_colour_brewer(palette = 'Set1')

lmod <- lm(yield ~ block + variety, data = oatvar)
summary(lmod)
anova(lmod)
anova(lm(yield ~ block, data = oatvar))
plot(lmod)
```

Comparison between complete randomized design (CRD) and randomized complete 
block design (RCBD):

Because `block` is significant, a model with blocking would be preferred

Can also consider efficiency

$\hat{\sigma}^2_{CRD}$ vs $\hat{\sigma}^2_{CRBD}$

```{r}
lmcrd <- lm(yield ~ variety, data = oatvar)
anova(lmcrd, lmod)

summary(lmcrd)$sig
summary(lmod)$sig

(summary(lmcrd)$sig / summary(lmod)$sig) ** 2
```

$\sim 66\%$ more observations are needed in CRD to get the same level of 
precision compared to CRBD

## Latin squares

Type of fractional factorial design for two blocking variables

Example: Compare 4 materials (which are our treatments): `A`, `B`, `C`, `D`, 
block 1: 4 positions `1`, `2`, `3`, `4`, block 2: 4 runs `1`, `2`, `3`, `4`, 
response: wear (weight of material worn)

```{r}
data(abrasion, package = 'faraway')
summary(abrasion)

xtabs(wear ~ run + position, data = abrasion)
matrix(abrasion$material, 4, 4, byrow = TRUE)
```

Note that each treatment is assigned to each block only once

The model:

$$y_{ijk} = \mu + \tau_i + \beta_j + \gamma_k + \epsilon_{ijk}$$

$i, j, k$ all have same number of levels $1, ..., t$, but not all appear in the 
data

* Latin square can be more efficient than RCBD
* block sizes (levels) should be equal to treatments (levels)
* can be used to compare three treatment factors
* cannot estimate interactions

```{r}
ggplot(abrasion) + 
  geom_line(aes(x = run, y = wear, colour = position, group = material))

ggplot(abrasion) + 
  geom_line(aes(x = position, y = wear, colour = run, group = material))

lmod <- lm(wear ~ material + run + position, data = abrasion)
summary(lmod)
drop1(lmod, test = 'F')
```