---
title: The MSC Memory Words Dataset
date: 'January 31, 2020'
author: John Koo
# output: beamer_presentation
# output: html_document
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE,
                      fig.lp = '')

options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

```{r}
import::from(magrittr, `%>%`)
```

# Midnight Scan Club Data Overview

* Most previous fMRI analyses involved looking at group effects (using 
group-aggregated data)
* Primary focus of MSC dataset: Consider individuals rather than groups

* 10 subjects 
* 5 tasks
* Includes additional information about each subject

# Incidental Memory Task (Words) Overview

* Subjects shown a sequence of words
* Words from the English Lexicon Project (Balota et al.)
    * Each session consisted of 24 words (40K in the ELP)
    * Subjects asked to categorize each word as abstract or concrete 
    (one of 13 semantic variables in the ELP)
* Each word presented for 1.7 seconds with 0.5 to 4.9 second intervals

# Visualizations

Using the `oro.nifti`[^1] and `neurobase`[^2] packages

[^1]: https://cran.r-project.org/web/packages/oro.nifti/index.html

[^2]: https://cran.rstudio.com/web/packages/neurobase/index.html

```{r}
home.dir <- '/N/project/clubneuro/MSC/fmriprep_ver1p3p2'

subjects <- 
  paste0('sub-MSC', stringr::str_pad(seq(10), 2, pad = '0', side = 'left'))
sessions <- 
  paste0('ses-func', stringr::str_pad(seq(10), 2, pad = '0', side = 'left'))

task <- 'memorywords'

# setwd(home.dir)
```

```{r}
setwd(file.path(home.dir, subjects[1], sessions[1], 'func'))

files <- dir() %>% 
  .[grepl(paste0('.+', task, '.+nii\\.gz'), .)]

x <- neurobase::readnii(files[5])
```

```{r}
oro.nifti::orthographic(x)
```

# Visualizations

```{r}
# oro.nifti::orthographic(x)
# neurobase::ortho2(x)
oro.nifti::slice(x, z = seq(dim(x)[3]))
```