---
title: 'S626'
subtitle: 'HW1'
author: 'John Koo'
output: pdf_document
# output: html_document
# geometry: "left=1cm,right=1cm,top=1cm,bottom=1.5cm"
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center', 
                      fig.height = 3, 
                      fig.width = 6, 
                      fig.dpi = 300)
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

# Problem 2

## Part a

Using $P(A) = \sum_i P(A \cap B_i)$

occupation | P(occupation)
-----------| -
farm       | `r .018 + .035 + .031 + .008 + .018`
operatives | `r .002 + .112 + .064 + .032 + .069`
craftsmen  | `r .001 + .066 + .094 + .032 + .084`
sales      | `r .001 + .018 + .019 + .01 + .051`
professional | `r .001 + .029 + .032 + .043 + .13`

## Part b

Similar to part (a)

occupation   | P(occupation)
-------------| -
farm         | `r .018 + .002 + .001 + .001 + .001`
operatives   | `r .035 + .112 + .066 + .018 + .029`
craftsmen    | `r .031 + .064 + .094 + .019 + .032`
sales        | `r .008 + .032 + .032 + .01 + .043`
professional | `r .018 + .069 + .084 + .051 + .13`

## Part c

Using $P(A | B) = \frac{P(A \cap B)}{P(B)}$

```{r echo = FALSE}
p_father_farm <- .018 + .035 + .031 + .008 + .018
p_son_farm <- .018 + .002 + .001 + .001 + .001
```

occupation   | P(son's occupation \| father is farmer)
-------------|-
farm         | `r round(.018 / p_father_farm, 3)`
operatives   | `r round(.035 / p_father_farm, 3)`
craftsmen    | `r round(.031 / p_father_farm, 3)`
sales        | `r round(.008 / p_father_farm, 3)`
professional | `r round(.018 / p_father_farm, 3)`

## Part d

Similar to part (c)

occupation   | P(father's occupation \| son is farmer)
-------------|-
farm         | `r round(.018 / p_son_farm, 3)`
operatives   | `r round(.002 / p_son_farm, 3)`
craftsmen    | `r round(.001 / p_son_farm, 3)`
sales        | `r round(.001 / p_son_farm, 3)`
professional | `r round(.001 / p_son_farm, 3)`

# Problem 3

## Part a

We can drop all the parts that do not depend on $x$

$p(x | y, z) = \frac{p(x, y, z)}{p(y, z)} \propto p(x, y, z)$ 
$\propto f(x, z) g(y, z) h(z) \propto f(x, z)$

## Part b

Similar to part (a), we can say:

$p(y | x, z) \propto p(x, y, z) \propto g(y, z)$

## Part c

$p(x, y | z) = \frac{p(x, y, z)}{p(z)} \propto p(x, y, z)$ 
$\propto f(x, z) g(y, z) h(z) \propto f(x, z) g(y, z)$ 
$\propto p(x, z) p(y, z)$
$= p(x | z) p(z) p(y | z) p(z)$
$\propto p(x | z) p(y | z)$

Suppose $p(x, y | z) \neq p(x | z) p(y | z)$, i.e., 
$p(x, y | z) = k p(x | z) p(y | z)$. Since $\int p(x, y | z) dx dy = 1$, 
$k = \big( \int p(x | z) p(y | z) dx dy \big)^{-1}$. But 
$\int p(x | z) p(y | z) dx dy = (\int p(x | z) dx) (\int p(y | z) dx)$
$= (1) (1) = 1 \implies k = 1$. Therefore, $p(x, y | z) = p(x | z) p(y | z)$.

# Problem 4

## Part a

$P_A(E) = \begin{cases} 
  0 & 6 \text{ is not the observed outcome} \\ 
  1 & 6 \text{ is the observed outcome}
\end{cases}$ 
since person $A$ already observed the outcome. 

On the other hand, let's say person $B$ assumes that the die is fair. Then 
$P_B(E) = 1/6$.

## Part b

If person $A$ is ignorant of soccer, then they may assign a uniform distribution 
across all teams. Of course, $A$ may not know how many teams compete, so a 
guess might be 200, which is approximately the number of countries in the world. 
So perhaps $P_A(E) = 1/200$.

On the other hand, if person $B$ is knowledgeable about soccer, then they may 
know that there are 32 teams that qualify for the World Cup and that Germany 
is almost guaranteed to qualify, so perhaps $P_B(E) = 1/32$. Person $B$ probably 
also knows that Germany won 2 out of 10 most recent World Cups, so perhaps 
$P_B(E) = 1/5$. 