---
title: "STAT-S675"
subtitle: 'Assignment 1'
author: "John Koo"
output: pdf_document
# output: html_document
---

# Section 1.5

## Exercise 1

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      fig.pos = 'H', 
                      fig.align = 'center')
```

```{r ex1, fig.width = 4, echo = FALSE}
pairs(dplyr::select(iris, 
                    `sepal length` = Sepal.Length, 
                    `sepal width` = Sepal.Width,
                    `petal length` = Petal.Length, 
                    `petal width` = Petal.Width), 
      main = 'Anderson\'s Iris Data', 
      pch = 21, 
      bg = c('red', 'green', 'blue')[unclass(iris$Species)])
```

* It appears that identifying setosas from the rest is possible with just either
petal length or petal width by setting a threshold value below which all points
will be classified as "setosa". 

* It appears that it's possible to come up with a reasonable method for
classifying versicolors and virginicas with a straight line (e.g., support 
vector machine) using any two measurements other than sepal length and width as 
a pair. Using petal length and petal width seems like it would result in the 
best model.

## Exercise 4

### Part a

#### In $\mathbb{R}^2$

WLOG we can set $\overrightarrow{x_1} = (0, 0)$, 
$\overrightarrow{x_2} = (2, 0)$, and $\overrightarrow{x_3} = (1, \sqrt{3})$. 
Then we have to find $\overrightarrow{x_4} = (x, y)$ that satisfies $\Delta$.

Our system of equations is then:

$$x^2 + y^2 = 1.21$$
$$(x - 2)^2 + y^2 = 1.21$$
$$(x - 1)^2 + (y - \sqrt{3})^2 = 1.21$$

From the first two equations:

$$x^2 + y^2 = (x - 2)^2 + y^2$$
$$x^2 = (x - 2)^2$$
$$x^2 = x^2 - 4x + 4$$
$$x = 1$$

Plugging this into the first equation from the system of equations yields:

$$1^2 + y^2 = 1.21$$
$$y = \pm \sqrt{.21}$$

So our possible solutions are $(1, \sqrt{.21})$ and $(1, -\sqrt{.21})$.

Finally, we can try both possiblities with our third equation:

$$(1 - 1)^2 + (\sqrt{.21} - \sqrt{3})^2 \approx 1.623 \neq 1.21$$
$$(1 - 1)^2 + (-\sqrt{.21} - \sqrt{3})^2 \approx 4.797 \neq 1.21$$

Therefore no combination of 4 vectors in $\mathbb{R}^2$ satisfies $\Delta$. 

#### In $\mathbb{R}^3$

Similar to the above, $\overrightarrow{x_1} = (0, 0, 0)$, 
$\overrightarrow{x_2} = (2, 0, 0)$, and 
$\overrightarrow{x_3} = (1, \sqrt{3}, 0)$. Then our system of equations is:

$$x^2 + y^2 + z^2 = 1.21$$
$$(x - 2)^2 + y^2 + z^2 = 1.21$$
$$(x - 1)^2 + (y - \sqrt{3})^2 + z^2 = 1.21$$

And we just need a nonempty solution space.

Similar to the $\mathbb{R}^2$ case, we can combine the first two equations to 
obtain $x = 1$. Plugging that into the first and third equations yields:

$$1^2 + y^2 + z^2 = 0^2 + (y - \sqrt{3})^2 + z^2$$
$$\implies 1 + y^2 = (y - \sqrt{3})^2$$

And solving for $y$ gives us $y = \frac{1}{\sqrt{3}}$.

Finally, we can plug $x$ and $y$ into either the first or second equation to 
obtain a value for $z$:

$$1 + \frac{1}{3} + z^2 = 1.21$$
$$z^2 = .21 - \frac{1}{3}$$

But this implies that $z^2 < 0$. Therefore no combination of 4 vectors in 
$\mathbb{R}^3$ satisfies $\Delta$.

### Part b

#### In $\mathbb{R}^2$

Similar to part a, we can start with some configuration of three vectors that 
satisfies part of $\Delta$ and then solve for the last vector. In this case, 
we can set $\overrightarrow{x_1} = (0, 0)$, $\overrightarrow{x_2} = (1, 0)$, 
$\overrightarrow{x_4} = (1, 1)$, and then solve for $\overrightarrow{x_3}$:

$$x^2 + y^2 = 4$$
$$(x - 1)^2 + y^2 = 3$$
$$(x - 1)^2 + (y - 1)^2 = 5$$

Combining the first two equations as we've done in the past problems yields
$x = 1$. Plugging that into the second equation yields $y = \pm \sqrt{3}$. 
Finally, we can try both $(1, \sqrt{3})$ and $(1, -\sqrt{3})$ with the third 
equation:

$$(1 - 1)^2 + (\sqrt{3} - 1)^2 \approx .536 \neq 5$$
$$(1 - 1)^2 + (-\sqrt{3} - 1)^2 \approx 7.464 \neq 5$$

Therefore there is no set of 4 vectors in $\mathbb{R}^2$ that works with 
$\Delta$.

#### In $\mathbb{R}^3$

To solve for $\overrightarrow{x_3} = (x, y, z)$:

$$x^2 + y^2 + z^2 = 4$$
$$(x - 1)^2 + y^2 + z^2 = 3$$
$$(x - 1)^2 + (y - 1)^2 + z^2 = 5$$

Combining the first two yields $x^2 - (x - 1)^2 = 1 \implies x = 1$.

Plugging $x = 1$ into the second and third equations and combining them yields 
$y^2 - (y - 1)^2 = -2 \implies y = -\frac{1}{2}$.

Plugging $x = 1$ and $y = -\frac{1}{2}$ into the second equation yields 
$z = \pm \frac{\sqrt{11}}{2}$. So we need to see if 
$(1, \frac{1}{2}, \pm \frac{\sqrt{11}}{2})$ is consistent with the first and 
third equations:

$$1 + \frac{1}{4} + \frac{11}{4} = 4$$
$$0 + \frac{1}{4} + \frac{11}{4} = 3$$

Therefore there are two sets of 4 vectors that works with $\Delta$.

## Exercise 5

We can generalize the methods used in exercise 4 for $\mathbb{R}^2$. 
Let $\Delta \in \mathbb{R} ^ {4 \times 4}$ be a dissimilarity matrix where 
$\delta{ij} = |\overrightarrow{x_i} - \overrightarrow{x_j}|_2$. Then we can 
start with $\overrightarrow{x_1} = (0, 0)$ and 
$\overrightarrow{x_2} = (\delta_{1, 2}, 0)$. Then to solve for 
$\overrightarrow{x_3} = (x, y)$ we have the system of equations: 

$$x^2 + y^2 = \delta_{1, 3}^2$$
$$(x - \delta_{1, 2})^2 + y^2 = \delta_{2, 3}^2$$

Assuming there is a solution, we end up with:

$$x = \frac{\delta_{1, 2}^2 + \delta_{1, 3}^2 - \delta_{2, 3}^2}{2\delta_{1, 2}}$$

$$y = \pm \sqrt{\delta_{1, 3}^2 - \frac{(\delta_{1, 2}^2 + \delta_{1, 3}^2 - \delta_{2, 3}^2)^2}{4\delta_{1, 2}^2}}$$

Due to symmetry, we only need to consider one of these two solutions. Note that 
for there to be a solution for $y \in \mathbb{R}$, the term under the suqare 
root needs to be positive.

Finally, we can let $\overrightarrow{x_4} = (u, v)$ and then solve:

$$u^2 + v^2 = \delta_{1, 4}^2$$
$$(u - \delta_{1, 2})^2 + v^2 = \delta_{2, 4}^2$$
$$(u - x)^2 + (v - y)^2 = \delta_{3, 4}^2$$

```{r ex_5}
G <- matrix(c(1, .1054, .0019, .0183, 
              .1054, 1, .0183, .0019, 
              .0019, .0183, 1, .1054, 
              .0183, .0019, .1054, 1), 
            nrow = 4, ncol = 4)
```

### Part a

```{r part_a}
Delta <- 1 - G

print(Delta)

x1 <- c(0, 0)
x2 <- c(Delta[1, 2], 0)
x3 <- c(
  (Delta[1, 2]^2 + Delta[1, 3]^2 - Delta[2, 3]^2) / 2 / Delta[1, 2], 
  sqrt(
    Delta[1, 3]^2 - 
      (Delta[1, 2]^2 + 
         Delta[1, 3]^2 - 
         Delta[2, 3]^2)^2 / 
      4 / Delta[1, 2]^2
    )
)
```

Let $\overrightarrow{x_1}$, $\overrightarrow{x_2}$, and $\overrightarrow{x_3}$ 
be defined as in the previous R chunk. We can verify that this works:

```{r part_a_check}
dist(rbind(x1, x2, x3))
```

Then we can use this to solve for $\overrightarrow{x_4}$ (boring algebra done 
elsewhere):

```{r solve_x4}
x4 <- 1:2
x4[1] <- (Delta[1, 4]^2 - Delta[2, 4]^2 + Delta[1, 2]^2) / 2 / Delta[1, 2]
x4[2] <- sqrt(Delta[1, 4]^2 - x4[1]^2)

# x4 has two solutions--calling the second one "x5"
x5 <- x4
x5[2] <- -x5[2]
```

And then we can check to see if either solution works:

```{r part_a_solution_check}
dist(rbind(x1, x2, x3, x4))
dist(rbind(x1, x2, x3, x5))
```

In either case, $\delta_{1, 4}$ and $\delta_{2, 4}$ work out (as it should since 
I solved for $\overrightarrow{x_4}$ using the first two) but $\delta{3, 4}$ is 
incorrect, implying that there is no configuration that works.

### Part b

We can do the same thing we did in part a:

```{r part_b}
Delta <- -log(G)

print(Delta)

x1 <- c(0, 0)
x2 <- c(Delta[1, 2], 0)
x3 <- c(
  (Delta[1, 2]^2 + Delta[1, 3]^2 - Delta[2, 3]^2) / 2 / Delta[1, 2], 
  sqrt(
    Delta[1, 3]^2 - 
      (Delta[1, 2]^2 + 
         Delta[1, 3]^2 - 
         Delta[2, 3]^2)^2 / 
      4 / Delta[1, 2]^2
    )
)
```

But in thise case, $\overrightarrow{x_3} \not\in \mathbb{R}^2$, so we can stop 
here.