---
title: "STAT-S675"
subtitle: 'Homework 4'
author: "John Koo"
date: "September 20, 2017"
output: 'pdf_document'
header-includes:
- \usepackage{float}
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      fig.pos = 'H', 
                      fig.align = 'center', 
                      dpi = 300, 
                      fig.height = 3, 
                      fig.width = 5, 
                      fig.retina = 1, 
                      cache = FALSE)
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

```{r packages}
import::from(readr, read_table)
library(ggplot2)
import::from(GGally, ggpairs)
theme_set(theme_bw())
import::from(magrittr, `%>%`)
import::from(ggrepel, geom_label_repel)
```

# Problem 1

```{r read_cork_data, warning = FALSE, message = FALSE}
cork.df <- read_table('http://pages.iu.edu/~mtrosset/Courses/675/mkb141.dat', 
                      col_names = c('north', 'east', 'south', 'west'))

ggpairs(cork.df)
```

The first thing we can notice is that all of these columns are highly 
correlated with each other. So we expect the first principal component to 
capture almost all of the variability.

```{r cork_pca}
cork.pca <- princomp(cork.df)
```

Here we want to use the covariance matrix (which it does by default) since 
all of the measurements are comparable to each other (i.e., same units, 
same type of measurement).

```{r cork_pca_barplot}
barplot(cork.pca$sdev ** 2, 
        ylab = 'variance')

barplot(cork.pca$sdev ** 2 / sum(cork.pca$sdev ** 2), 
        ylab = 'relative variance')

cork.pca$sdev[1] ** 2 / sum(cork.pca$sdev ** 2)
```

The first principal component captures >90\% of the variance.

```{r cork_loadings}
cork.pca$loadings
```

Here we can see the actual components. One thing to note is that the signs are 
arbitrary as long as they're consistent (e.g., all positive is equivalent to all 
negative). An interpretation of this might be something like:

* The first component has the same sign and roughly similar magnitudes for 
all of north, east, south, and west, so we can interpret it as the overall 
magnitude of all of the variables. This is consistent with our observation 
that all of the columns are highly correlated with each other.

* The second component has positive values for north and east while having 
negative values for south and west. This can be interpreted as measuring how 
different north and east are from south and west. 

* The third and fourth have similar interpretations as the second. The third 
measures how different north and west are from east and south, while the fourth 
measures how different north and south are from east and west.

* One quick observation is that the second principal component measures 
north vs south and east vs west, which agrees somewhat with our intuition. 
The sun is more to the north or to the south depending on the season and 
hemisphere, so it's reasonable to hypothesize that there is some natural 
difference there. Similarly, the sun rises in the east and sets in the west, 
and it's reasonable to hypothesize that there would be some difference there as 
well.

```{r cork_biplot, fig.height = 5}
biplot(cork.pca)
```

From the biplot, we can see that the angles between each pair of columns is 
very close, implying strong correlation. The two columns with the largest 
angle between them are east and west. This is consistent with our observation 
from the matrix scatterplot showing that these two columns are the least
correlated.

```{r cork_scatter}
cork.df %>% 
  dplyr::mutate(row = rownames(.)) %>% 
  ggplot() + 
  geom_point(aes(x = east, y = west)) + 
  stat_smooth(aes(x = east, y = west), method = 'lm') + 
  geom_label_repel(aes(x = east, y = west, label = row))
```

From eyeballing the biplot, we can see how far off from a linear fit of east vs 
west each point is. For example, the 8th point in the biplot has a very small 
yet positive "east" component while it has a very negative "west" component, and 
in the scatterplot of west vs east the 8th point is very far from the linear 
fit line. On the other hand, the 15th point has relatively similar values 
for east and west in the biplot and in the scatterplot it lies very close to the 
linear fit line.

# Problem 2

Let $q = 30000$. The following R commands generate a random $q \times 10$ 
matrix:

```{r p2_setup}
set.seed(675)
q <- 30000
l <- 10
Q <- matrix(rnorm(q*l),ncol=l)/sqrt(q)
```

Each column of `Q` is a vector in $\mathbb{R}^q$.

Investigate how nearly orthonormal these vectors turn out to be.

Two vectors are orthonormal iff they are unit vectors (i.e., length 1) and they 
are orthogonal (i.e., their cross product is 0). 

## Vector length

We can compute the length of each of these vectors:

```{r p2_length}
vector.lengths <- sqrt(colSums(Q ** 2))
sort(vector.lengths)
summary(vector.lengths)
ggplot() + 
  geom_histogram(aes(x = vector.lengths), 
                 bins = 6, colour = 'black', fill = 'white')
```

Here we can see that all of the lengths are very close to 1. This is expected 
since we know that for each vector, the elements are normally distributed 
with $\mu = 0$ and $\sigma = \frac{1}{\sqrt{q}}$, so 
$E[|x_i|_2] = E[\sqrt{\sum_j x_{ij}^2}] = \sqrt{q \sigma^2} = 1$ 
(using some very loose, hand-wavy logic). So on average, the vectors will be 
of length 1. 

## Orthogonality

Note that $[Q^T Q]_{ij} = \sum_k Q_{ki} Q_{kj} = <Q_{\cdot, i}, Q_{\cdot, j}>$

So we can just compute $Q^T Q$ and look at either the upper or lower triangular 
terms.

```{r p2_orthogonality}
QTQ <- t(Q) %*% Q

inner.products <- QTQ[lower.tri(QTQ)]
summary(inner.products)
ggplot() + 
  geom_histogram(aes(x = inner.products), 
                 bins = 8, colour = 'black', fill = 'white')
```

Here we can see that the inner products are all close to 0. Using similar 
hand-wavy logic as before, we can say something like: 

$$E[<Q_{\cdot, i}, Q_{\cdot, j}>] 
= E[\sum_k Q_{ki} Q_{kj}] 
\sim q \times cov(Q_{\cdot, i}, Q_{\cdot, j}) = 0$$

since the columns of $Q$ are independent. 

# Problem 3

Download the $201 \times 2$ data matrix `X.spiral` and scan it into R. Use the R 
Package Installer to download and install the package `pracma`. Let `q = 30000`. 
The following R commands construct an isometrically equivalent configuration of 
$n = 201$ points in $\mathbb{R}^q$:

```{r p3_setup}
X.spiral <- read.table('http://pages.iu.edu/~mtrosset/Courses/675/X.spiral') %>% 
  as.matrix()
library(pracma)
q <- 30000
A <- matrix(rnorm(q*2),ncol=2)
A.QR <- gramSchmidt(A)
Y <- X.spiral %*% t(A.QR$Q)
```

# Part a

Vectors containing the $n(n - 1)/2$ pairwise Euclidean distances of each 
configuration can be computed using the `dist` function. Verify that the 
configurations `Y` and `X.spiral` are isometrically equivalent by showing that 
they have the same pairwise distances, e.g., by showing that the *root mean 
squared error* (RMSE) vanishes:

```{r p3_setup_a}
dX <- dist(X.spiral)
dY <- dist(Y)
m <- length(dX)
rmse <- sqrt(sum((dY-dX)^2)/m)

# compare to:
.Machine$double.eps
```

The RMSE here is $`r signif(rmse, 4)`$. 

# Part b

Investigate the extent of distortion introduced by the following random 
projection technique, which replaces the $n \times q$ data matrix `Y` with the 
$n \times 10$ data matrix `Y10`:

```{r p3_setup_b}
Q10 <- matrix(rnorm(q*10),ncol=10)/sqrt(q)
Y10 <- Y %*% Q10
```

Find the distance matrix of `Y10` and compute its RMSE to the distance matrix 
of `Y`:

```{r p3_b_dist}
dY10 <- dist(Y10)

sqrt(sum((dY - dY10) ** 2) / m)
```

Here the RMSE is quite large, especially when compared to the typical spread of 
the values of the distance matrix of Y:

```{r p3_b_dist_normalized}
sqrt(sum((dY - dY10) ** 2) / m) / sd(dY)
```

So the RMSE is twice the standard deviation of the distances in the original Y. 

We can also see how this changes as the number of random projections changes:

```{r random_projections}
d <- seq(10, 1000, 10)
rmse.vector <- sapply(d, function(x) {
  Qd <- matrix(rnorm(q * x), ncol = x) / sqrt(q)
  Yd <- Y %*% Qd
  dYd <- dist(Yd)
  sqrt(sum((dY - dYd) ** 2) / m)
})

ggplot() + 
  geom_point(aes(x = d, y = rmse.vector)) + 
  labs(y = 'RMSE')
```

And as expected, the RMSE decreases as we increase the number of projections. 