---
title: "STAT-S675"
subtitle: 'Homework 8'
author: "John Koo"
output: pdf_document
# output: html_document
urlcolor: blue
header-includes:
- \usepackage{float}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center')
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

[Link to assignment](http://pages.iu.edu/~mtrosset/Courses/675/hwManifold.txt)

```{r setup_for_assignment}
dp <- loadNamespace('dplyr')
import::from(magrittr, `%>%`, `%<>%`)
import::from(foreach, foreach, `%do%`)
import::from(readr, read_table2)
import::from(scatterplot3d, scatterplot3d)
library(ggplot2)

source('http://pages.iu.edu/~mtrosset/Courses/675/out.r')
source('http://pages.iu.edu/~mtrosset/Courses/675/stress.r')
source('http://pages.iu.edu/~mtrosset/Courses/675/manifold.r')

theme_set(theme_bw())
```

# Exercise 7.4.1

[From the text]

Here we will construct $\Delta$ from the circle (since the result is identical 
in either case). 

```{r construct_delta, cache = TRUE}
# number of points
n <- 200

# radius of circle
r <- 1 / pi

# construct dissimilarity matrix
theta <- seq(2 * pi / n, 2 * pi, 2 * pi / n)
Delta <- foreach(i = theta, .combine = rbind) %do% {
  foreach(j = theta, .combine = c) %do% {
    theta.ij <- abs(i - j)
    if (theta.ij > pi) theta.ij <- 2 * pi - theta.ij
    arc.length <- theta.ij * r
    return(arc.length)
  }
}
```

```{r embed_delta}
# spectral decomposition of tau(Delta2)
Delta2 <- Delta ** 2
B <- mds.tau(Delta2)
B.eigen <- eigen(B)

# find first two components
projection.df <- B.eigen$vectors[, 1:2] %>% 
  apply(2, function(x) x * sqrt(B.eigen$values[1:2])) %>% 
  as.data.frame()

# plot first two components
ggplot(projection.df) + 
  geom_point(aes(x = V1, y = V2)) + 
  coord_fixed()
```

We can look at the rows of $B = \tau(\Delta_2)$ to obtain $b^2$:

```{r obtain_b}
summary(diag(B))
b <- sqrt(mean(diag(B)))
print(b ** 2)
```

Then $b^2 \approx `r round(b ** 2, 4)`$ (and $b \approx `r round(b, 4)`$).

If $b$ is the radius of the embedding, this is consistent with our intuition. 
The longest arc length in $S$ is 1, so we expect the longest pairwise Euclidean 
distance of the embedding, $2b$, or the diameter of of the embedding, to be 
approximately this length. 

To find $a$, the angle between two consecutive points, we want $A_{i, i + 1}$, 
where $A = \cos^{-1}\frac{B}{b^2}$.

```{r compute_a}
A <- acos(B / b ** 2)
print(A[1:6, 1:6])
```

We can see that $a \approx 0.02449$. In the original circle $S$, the angle 
between two consecutive values was $\frac{2 \pi}{200} \approx 0.0314$.

### Analytical derivation of $b^2$

Define $B = \tau(\Delta_2) = -\frac{1}{2} P \Delta_2 P$ where 
$P = I - \frac{e e^{\top}}{n}$. Then:

$$B = \tau(\Delta_2) = -\frac{1}{2} P \Delta_2 P$$
$$= -\frac{1}{2} (I - \frac{e e^{\top}}{n}) \Delta_2 (I - \frac{e e^{\top}}{n})$$
$$= -0.5 (\Delta_2 - \frac{e e^{\top}}{n} \Delta_2) (I - \frac{e e^{\top}}{n})$$
$$= -0.5 (\Delta_2 - \Delta_2 \frac{e e^\top}{n} - 
\frac{e e^\top}{n} \Delta_2 + \frac{e e^\top e e^\top}{n^2} \Delta_2)$$

We know that $e e^\top = n$ and since $\Delta_2$ and $\frac{e e^\top}{n}$ are 
symmetric, the two matrices commute. Therefore, this is equal to:

$$= -0.5 (\Delta_2 - 2 \frac{e e^\top}{n} \Delta_2
+ \frac{n e e^\top}{n^2} \Delta_2)$$
$$= -0.5 (\Delta_2 - \frac{e e^\top}{n} \Delta_2)$$
$$= -\frac{1}{2} (I - \frac{e e^\top}{n}) \Delta_2$$
$$= -\frac{1}{2} P \Delta_2$$
We know that the diagonal entries of this matrix are just half the row means of 
$\Delta_2$ subtracted by the diagonal entries of $\Delta_2$. However, 
$diag(\Delta_2) = 0$, so ...

$$[B]_{jj} = diag(\frac{1}{2} P \Delta_2) = \frac{1}{2n} \sum_k^n \delta_{kj}^2$$

This is equivalent $\forall j \leq n$, so we can just look at the first row of 
$\Delta_2$. 
$\delta_{1 \cdot}^2 = [0, .01^2, 0.2^2, 0.3^2, ..., .99^2, 1, .99^2, ..., .01^2]$ 
$= \frac{1}{100^2} [0, 1, 4, 9, ..., 9801, 10000, 9801, ..., 4, 1]$ 

Then $\sum_k^n \delta_{kj}^2 = \frac{1}{100^2} (\sum_k^{100} k^2 + \sum_k^{99} k^2)$ 
$= \frac{1}{100^2}(\frac{(100)(101)(201)}{6} + \frac{(99)(100)(199)}{6})$ 
$= 66.67$

Therefore, $[B]_{jj} = diag(\frac{1}{2} P \Delta_2) = 
\frac{1}{2n} \sum_k^n \delta_{kj}^2 = \frac{66.67}{400} = 0.166675$

# Other problems

```{r load_data}
file.url <- 'http://pages.iu.edu/~mtrosset/Courses/675/manifold.y'
Y <- read_table2(file.url, col_names = FALSE)

scatterplot3d(Y$X1, Y$X2, Y$X3)
```

# Problem 1

CMDS

```{r p1}
# CMDS on Y
Y.dist <- mds.edm1(Y)
Y.cmds <- cmdscale(Y.dist)

# plot first two components
Y.cmds.df <- Y.cmds %>% 
  as.data.frame() %>% 
  dp$mutate(id = rownames(.))
ggplot(Y.cmds.df) + 
  geom_text(aes(x = V1, y = V2, label = id)) + 
  coord_fixed()

# rank by first component
rank.1 <- rank(Y.cmds.df$V1)
print(rank.1)
```

# Problem 2

Isomap with EDM-1 weights

```{r p2}
# set k
k <- 4

# dimension of Y
n <- nrow(Y)

# define dissimilarity matrix for Y
Y.knn <- graph.knn(Y.dist, k)
Y.dis <- graph.dis(Y.knn, Y.dist)
Y.shortest <- graph.short(Y.dis)

Y.isomap.df <- Y.shortest %>% 
  cmdscale() %>% 
  as.data.frame() %>% 
  dp$mutate(id = rownames(.))

ggplot(Y.isomap.df) + 
  geom_text(aes(x = V1, y = V2, label = id)) + 
  coord_fixed()

rank.2 <- rank(Y.isomap.df$V1)
print(rank.2)
```

# Problem 3

Isomap with unit weights

```{r p3}
Y.dis.unit <- graph.dis(Y.knn, matrix(rep(1, n ** 2), ncol = n, nrow = n))
Y.shortest.unit <- graph.short(Y.dis.unit)

Y.isomap.unit.df <- Y.shortest.unit %>% 
  cmdscale() %>% 
  as.data.frame() %>% 
  dp$mutate(id = rownames(.))

ggplot(Y.isomap.unit.df) + 
  geom_text(aes(x = V1, y = V2, label = id)) + 
  coord_fixed()

rank.3 <- rank(Y.isomap.unit.df$V1, ties.method = 'first')
print(rank.3)
```

# Problem 4

Laplacian eigenmap with adjacency matrix

```{r p4}
Y.adj <- Y %>% 
  mds.edm1() %>% 
  graph.knn(k) %>% 
  graph.adj()

L.adj <- graph.laplacian(Y.adj)
L.adj.eigen <- eigen(L.adj)
L.adj.df <- cbind(
  L.adj.eigen$vectors[, n - 1] / sqrt(L.adj.eigen$values[n - 1]),
  L.adj.eigen$vectors[, n - 2] / sqrt(L.adj.eigen$values[n - 2])
) %>%
  as.data.frame() %>%
    dp$mutate(id = as.numeric(rownames(.)))

ggplot(L.adj.df) + 
  geom_text(aes(x = V1, y = V2, label = id)) + 
  coord_fixed()

rank.4 <- rank(L.adj.df$V1)
print(rank.4)
```

# Problem 5

Laplacian eigenmap with heat kernel

```{r p5}
s <- 5

Y.heat <- Y %>% 
  mds.edm1() %>% 
  graph.heat(s)

L.heat <- graph.laplacian(Y.heat)

L.heat.eigen <- eigen(L.heat)

L.heat.df <- cbind(
  L.heat.eigen$vectors[, n - 1] / sqrt(L.heat.eigen$values[n - 1]),
  L.heat.eigen$vectors[, n - 2] / sqrt(L.heat.eigen$values[n - 2])
) %>%
  as.data.frame() %>%
    dp$mutate(id = as.numeric(rownames(.)))

ggplot(L.heat.df) + 
  geom_text(aes(x = V1, y = V2, label = id)) + 
  coord_fixed()

rank.5 <- rank(L.heat.df$V1)
print(rank.5)
```

# Problem 6

```{r p6}
cor(cbind(seq(n), rank.1, rank.2, rank.3, rank.4, rank.5), 
    method = 'spearman')
```

We can see that the CMDS method is the odd one out here. The first component of 
the CMDS method fails to recover any structure. 

The Isomap and various Laplacian eigenmap methods perform better and all have 
similar results, with the heat map performing the best, at least based on just 
the ranking of the first component. 

The fact that CMDS failed here matches our intuition. CMDS requires some sort of 
correlation and cannot detect pairwise similarities very well. Since our 
original data `Y` did not have this correlated structure, it failed to find a 
good embedding. 