---
title: "STAT-S675"
subtitle: 'Homework 8'
author: "John Koo"
# output: pdf_document
output: html_document
urlcolor: blue
header-includes:
- \usepackage{float}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = NA, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.pos = 'H', 
                      fig.align = 'center')
options(xtable.comment = FALSE, 
        xtable.table.placement = 'H')
```

[Link to assignment](http://pages.iu.edu/~mtrosset/Courses/675/hwCluster.txt)

# Exercise 8.5.1

Let the points $\{-1.1, -1, -.9, -.1, 0, .1, .9, 1, 1.1)\} \subset \mathbb{R}$ 
be grouped into three clusters. Then a natural clustering of these nine points might be to have the first three points in one cluster, the middle three points in another cluster, and the last three points in the last cluster.

```{r cluster_setup}
library(ggplot2)
theme_set(theme_bw())

cluster.df <- dplyr::data_frame(
  x = c(-1.1, -1, -.9, -.1, 0, .1, .9, 1, 1.1), 
  cluster = rep(c('1', '2', '3'), each = 3)
)

ggplot(cluster.df) + 
  coord_fixed() + 
  geom_hline(yintercept = 0) + 
  scale_colour_brewer(palette = 'Set1') + 
  geom_point(aes(x = x, y = 0, colour = cluster)) + 
  labs(x = NULL, y = NULL)
```

Then $ES_{12} = .8 = ES_{23}$ and $ES_{13} = 1.8$. But $1.8 > .8 + .8$. 
Therefore, single-linkage doesn't satisfy the triangle inequality and so is not 
EDM-1.

# Exercise 8.5.2

# Exercise 8.5.3

```{r p3_setup}
r <- seq(11)
x <- c((r - 1) / 10, 2)

ggplot() + 
  geom_hline(yintercept = 0) + 
  coord_fixed() + 
  geom_point(aes(x = x, y = 0)) + 
  labs(x = NULL, y = NULL)
```

If we put the 11 leftmost points in one cluster and the point at $x = 2$ in 
another:

```{r cluster_1}
sum(as.matrix(dist(x[seq(11)])))
```

On the other hand, if we put the two outermost points in one cluster and the 10 
points between them in another cluster:

```{r cluster_2}
sum(as.matrix(dist(x[seq(2, 11)]))) + sum(as.matrix(dist(x[c(1, 12)])))
```

## Optimal clustering with $K = 2$

We can see that the optimal clustering would divide the left $n$ points into one 
cluster and the right $12 - n$ points into another cluster. So now we just have 
to find $n$. 

```{r optimal clustering}
W.vector <- sapply(seq(2, 11), function(n) {
  sum(as.matrix(dist(x[seq(n)]))) + sum(as.matrix(dist(x[seq(n, 12)])))
})

ggplot() + 
  geom_point(aes(x = seq(2, 11), y = W.vector)) + 
  labs(x = 'n', y = 'W') + 
  scale_x_continuous(breaks = seq(12))
```

So $W$ is minimized when $n = 7$, i.e., when the seven left-most points are in 
one cluster and the 5 right-most points are in the other cluster.